---
title: "Barcelona El Prat Outbound Flights and Rail Alternatives"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

This notebook audits the corrected route and airport files in this repo,
extracts all outbound routes from Barcelona El Prat (`LEBL`), maps those flight
links, and then estimates a rail-based shortest path for each destination using
the GL2017 European rail network.

The rail workflow is deliberately conservative and reproducible:

- airports are snapped to their nearest passenger-capable rail segment
- the GL2017 rail network is split at line intersections before routing
- travel time is the sum of:
  - substantial contiguous high-speed segment distance / `HSR_SPEED_KMH`
  - remaining routed rail distance / `CONVENTIONAL_SPEED_KMH`
  - airport-to-rail connector distance at `CONNECTOR_SPEED_KMH` on both ends
  - a fixed transfer penalty of `TRANSFER_PENALTY_H`
- routes with no connected rail path remain `NA`
- airports farther than `MAX_CONNECTOR_KM` from passenger rail are treated as
  not rail-feasible
- only contiguous HSR runs of at least `MIN_CONTIGUOUS_HSR_KM` are counted as
  effective high-speed rail in the travel-time calculation

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 13,
  fig.height = 8
)
```

## Packages and Paths

```{r packages}
suppressPackageStartupMessages({
  library(sf)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(purrr)
  library(stringr)
  library(ggplot2)
  library(scales)
  library(ggrepel)
  library(here)
  library(sfnetworks)
  library(tidygraph)
  library(units)
})

source(here::here("scripts", "paths.R"))
```

```{r constants}
ANALYSIS_CRS <- 3035
PASSENGER_RAIL_CLASSES <- c("Passenger", "Passenger and freight")
HSR_SPEED_KMH <- 250
CONVENTIONAL_SPEED_KMH <- 110
CONNECTOR_SPEED_KMH <- 45
MAX_CONNECTOR_KM <- 25
MIN_CONTIGUOUS_HSR_KM <- 50
TRANSFER_PENALTY_H <- 0.5
ORIGIN_ICAO <- "LEBL"

routes_path <- file.path(data_clean, "master_routes_clean.csv")
airports_path <- file.path(data_clean, "all_airports.gpkg")
rail_path <- file.path(data_raw, "shapefiles", "railways_GL2017_EU.shp")
countries_path <- file.path(data_clean, "ne_10m_countries.gpkg")

required_paths <- c(routes_path, airports_path, rail_path, countries_path)
stopifnot(all(file.exists(required_paths)))

dir.create(outputs, recursive = TRUE, showWarnings = FALSE)
```

## File Audit

```{r read-inputs}
routes_raw <- read_csv(routes_path, show_col_types = FALSE)

airports_raw <- st_read(airports_path, quiet = TRUE) %>%
  st_make_valid() %>%
  st_transform(ANALYSIS_CRS)

rail_raw <- st_read(rail_path, quiet = TRUE) %>%
  st_make_valid() %>%
  st_transform(ANALYSIS_CRS)

countries <- st_read(countries_path, quiet = TRUE) %>%
  st_make_valid() %>%
  st_transform(ANALYSIS_CRS)
```

```{r column-audit}
cat("master_routes_clean.csv columns:\n")
print(names(routes_raw))

cat("\nall_airports.gpkg columns:\n")
print(names(airports_raw))

cat("\nrailways_GL2017_EU.shp columns:\n")
print(names(rail_raw))
```

```{r audit-summary}
cat("Rows in master_routes_clean.csv:", nrow(routes_raw), "\n")
cat("Rows in all_airports.gpkg:", nrow(airports_raw), "\n")
cat("Rows in GL2017 rail shapefile:", nrow(rail_raw), "\n")

cat("\nDistinct start airports named Barcelona / LEBL in route table:\n")
print(
  routes_raw %>%
    filter(start_icao == ORIGIN_ICAO | str_detect(start_airport, regex("Barcelona", ignore_case = TRUE))) %>%
    distinct(start_airport, start_icao)
)
```

## Barcelona El Prat Outbound Routes

The corrected route table is already close to one row per route, but I still
collapse to one record per `end_icao` as a safeguard. For any duplicated
destination ICAO, the notebook keeps the maximum observed route passenger count
for plotting.

```{r airport-cleaning}
airports_clean <- airports_raw %>%
  mutate(
    station_like = str_detect(
      str_to_upper(AirportName),
      "RAILWAY STATION|\\bHBF\\b|\\bSTATION\\b"
    ),
    name_length = nchar(AirportName)
  ) %>%
  arrange(ICAO, station_like, desc(name_length)) %>%
  group_by(ICAO) %>%
  slice(1) %>%
  ungroup() %>%
  select(-name_length)

safe_max <- function(x) {
  if (all(is.na(x))) {
    return(NA_real_)
  }
  max(x, na.rm = TRUE)
}

barcelona_routes <- routes_raw %>%
  filter(start_icao == ORIGIN_ICAO) %>%
  group_by(end_icao) %>%
  summarise(
    total_flights = safe_max(total_flights),
    route_passengers = safe_max(route_passengers),
    raw_rows = n(),
    passenger_sum_raw = sum(route_passengers, na.rm = TRUE),
    passenger_max_raw = safe_max(route_passengers),
    .groups = "drop"
  ) %>%
  left_join(
    airports_clean %>%
      st_drop_geometry() %>%
      transmute(
        end_icao = ICAO,
        end_airport = AirportName,
        end_iata = IATA,
        end_city = City_Name,
        end_country = Country_Name
      ),
    by = "end_icao"
  ) %>%
  filter(!is.na(end_airport)) %>%
  arrange(desc(route_passengers))

barcelona_airport <- airports_clean %>%
  filter(ICAO == ORIGIN_ICAO)

destination_airports <- airports_clean %>%
  filter(ICAO %in% barcelona_routes$end_icao) %>%
  transmute(
    end_icao = ICAO,
    end_airport_geom = AirportName
  )

flight_destinations <- destination_airports %>%
  left_join(barcelona_routes, by = "end_icao") %>%
  filter(!is.na(route_passengers))

cat("Barcelona outbound rows in source table:", nrow(routes_raw %>% filter(start_icao == ORIGIN_ICAO)), "\n")
cat("Distinct Barcelona destination ICAOs after collapse:", nrow(barcelona_routes), "\n")
cat("Routes with non-missing passenger counts:", sum(!is.na(barcelona_routes$route_passengers)), "\n")
```

```{r duplicate-diagnostics}
duplicate_routes <- routes_raw %>%
  filter(start_icao == ORIGIN_ICAO) %>%
  count(end_icao, sort = TRUE) %>%
  filter(n > 1)

duplicate_airports <- airports_raw %>%
  st_drop_geometry() %>%
  count(ICAO, sort = TRUE) %>%
  filter(n > 1)

duplicate_routes
duplicate_airports
```

## Flight Map from El Prat

```{r arc-helpers}
make_arc <- function(from_xy, to_xy, curvature = 0.035, n = 120) {
  t <- seq(0, 1, length.out = n)
  base <- (1 - t) %o% from_xy + t %o% to_xy
  direction <- to_xy - from_xy
  seg_length <- sqrt(sum(direction^2))

  if (seg_length == 0) {
    return(st_linestring(base))
  }

  unit_perp <- c(-direction[2], direction[1]) / seg_length
  curve_side <- ifelse(to_xy[1] >= from_xy[1], 1, -1)
  offset <- curve_side * curvature * seg_length * sin(pi * t)
  curve <- base + offset %o% unit_perp
  st_linestring(curve)
}

origin_xy <- st_coordinates(barcelona_airport)[1, 1:2]
dest_xy <- st_coordinates(flight_destinations)[, 1:2, drop = FALSE]

flight_arc_sfc <- st_sfc(
  lapply(seq_len(nrow(flight_destinations)), function(i) {
    make_arc(origin_xy, dest_xy[i, ], curvature = 0.035)
  }),
  crs = ANALYSIS_CRS
)

flight_arcs <- st_sf(
  flight_destinations %>%
    st_drop_geometry() %>%
    select(end_icao, end_airport, route_passengers, raw_rows),
  geometry = flight_arc_sfc
)
```

```{r map-extent}
map_bbox <- st_bbox(
  st_union(
    st_geometry(barcelona_airport),
    st_geometry(destination_airports)
  )
)

x_pad <- (map_bbox$xmax - map_bbox$xmin) * 0.10
y_pad <- (map_bbox$ymax - map_bbox$ymin) * 0.10

map_xlim <- c(map_bbox$xmin - x_pad, map_bbox$xmax + x_pad)
map_ylim <- c(map_bbox$ymin - y_pad, map_bbox$ymax + y_pad)
```

```{r flight-map, fig.cap="Barcelona El Prat outbound flights, coloured by passengers"}
top_labels <- barcelona_routes %>%
  filter(!is.na(route_passengers)) %>%
  slice_max(order_by = route_passengers, n = 15, with_ties = FALSE) %>%
  select(end_icao)

flight_map <- ggplot() +
  geom_sf(data = countries, fill = "#F8F8F6", color = "#C9CDD1", linewidth = 0.18) +
  geom_sf(
    data = flight_arcs,
    aes(color = route_passengers, linewidth = route_passengers),
    alpha = 0.78,
    lineend = "round",
    show.legend = "line"
  ) +
  geom_sf(
    data = flight_destinations,
    aes(size = route_passengers),
    shape = 21,
    fill = "white",
    color = "#19324A",
    stroke = 0.35,
    alpha = 0.95
  ) +
  geom_sf(
    data = barcelona_airport,
    shape = 21,
    size = 5,
    fill = "#F25F5C",
    color = "#5C1A18",
    stroke = 0.8
  ) +
  geom_text_repel(
    data = flight_destinations %>%
      filter(end_icao %in% top_labels$end_icao) %>%
      cbind(st_coordinates(.)),
    aes(X, Y, label = end_airport),
    size = 3.1,
    box.padding = 0.35,
    point.padding = 0.2,
    min.segment.length = 0
  ) +
  scale_color_viridis_c(
    option = "plasma",
    trans = "log10",
    labels = label_number(big.mark = ","),
    na.value = "#B8C4CE",
    name = "Route passengers"
  ) +
  scale_linewidth_continuous(range = c(0.18, 1.5), guide = "none") +
  scale_size_continuous(range = c(1.1, 5.2), labels = label_number(big.mark = ","), guide = "none") +
  coord_sf(xlim = map_xlim, ylim = map_ylim, expand = FALSE) +
  labs(
    title = "Outbound Flight Routes from Barcelona El Prat",
    subtitle = paste0(
      sum(!is.na(barcelona_routes$route_passengers)),
      " outbound routes with observed passenger counts to ",
      nrow(barcelona_routes),
      " destination ICAOs"
    ),
    caption = "Source files: data/clean/master_routes_clean.csv and data/clean/all_airports.gpkg"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.background = element_rect(fill = "#EAF3FB", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "#D7E0EA", linewidth = 0.22),
    panel.grid.minor = element_blank(),
    axis.title = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )

flight_map
```

```{r save-flight-map, eval=FALSE}
ggsave(
  file.path(outputs, "barcelona_el_prat_flight_routes.png"),
  flight_map,
  width = 13,
  height = 8,
  dpi = 300,
  bg = "white"
)
```

## Rail Network Routing

```{r prepare-rail}
rail_passenger <- rail_raw %>%
  filter(RAILWAYS_A %in% PASSENGER_RAIL_CLASSES) %>%
  filter(!st_is_empty(geometry), !is.na(geometry)) %>%
  st_collection_extract("LINESTRING") %>%
  mutate(
    speed_kmh = if_else(TYPE == "High speed", HSR_SPEED_KMH, CONVENTIONAL_SPEED_KMH),
    rail_class = if_else(TYPE == "High speed", "High speed", "Conventional")
  )

cat("Passenger-capable GL2017 rail segments retained:", nrow(rail_passenger), "\n")
print(rail_passenger %>% st_drop_geometry() %>% count(TYPE, sort = TRUE))
```

```{r snap-airports-to-rail}
route_airports <- airports_clean %>%
  filter(ICAO %in% c(ORIGIN_ICAO, barcelona_routes$end_icao)) %>%
  transmute(ICAO, AirportName)

nearest_edge_idx <- st_nearest_feature(route_airports, rail_passenger)

snapped_geoms <- lapply(seq_len(nrow(route_airports)), function(i) {
  pair <- st_nearest_points(route_airports[i, ], rail_passenger[nearest_edge_idx[i], ])
  snapped_pair_points <- st_cast(pair, "POINT")
  snapped_pair_points[2]
})

snapped_sfc <- do.call(c, snapped_geoms)
st_crs(snapped_sfc) <- st_crs(ANALYSIS_CRS)

snapped_airports <- st_as_sf(
  route_airports %>% st_drop_geometry(),
  geometry = snapped_sfc
) %>%
  mutate(
    nearest_edge_idx = nearest_edge_idx,
    access_km = as.numeric(
      set_units(st_distance(route_airports, geometry, by_element = TRUE), "km")
    ),
    access_time_h = access_km / CONNECTOR_SPEED_KMH,
    within_connector_limit = access_km <= MAX_CONNECTOR_KM
  )

snapped_airports %>%
  st_drop_geometry() %>%
  slice_max(order_by = access_km, n = 12)
```

```{r build-network}
rail_net <- rail_passenger %>%
  as_sfnetwork(directed = FALSE) %>%
  convert(to_spatial_subdivision)

rail_net <- st_network_blend(rail_net, snapped_airports)

rail_net <- rail_net %>%
  activate("edges") %>%
  mutate(
    edge_length_km = as.numeric(set_units(st_length(geometry), "km")),
    edge_time_h = edge_length_km / speed_kmh
  )

network_nodes <- st_as_sf(rail_net, "nodes")
network_edges <- st_as_sf(rail_net, "edges")

snapped_airports <- snapped_airports %>%
  mutate(node_id = st_nearest_feature(geometry, network_nodes))
```

```{r route-function}
origin_node_id <- snapped_airports %>%
  filter(ICAO == ORIGIN_ICAO) %>%
  pull(node_id)

origin_access_h <- snapped_airports %>%
  filter(ICAO == ORIGIN_ICAO) %>%
  pull(access_time_h)

origin_access_km <- snapped_airports %>%
  filter(ICAO == ORIGIN_ICAO) %>%
  pull(access_km)

build_rail_route <- function(dest_code) {
  dest_info <- snapped_airports %>%
    filter(ICAO == dest_code)

  if (!isTRUE(dest_info$within_connector_limit) || !isTRUE(snapped_airports %>% filter(ICAO == ORIGIN_ICAO) %>% pull(within_connector_limit))) {
    return(
      st_sf(
        end_icao = dest_code,
        rail_connected = FALSE,
        connector_feasible = FALSE,
        rail_km = NA_real_,
        high_speed_km = NA_real_,
        raw_high_speed_km = NA_real_,
        conventional_km = NA_real_,
        rail_in_vehicle_h = NA_real_,
        origin_access_km = origin_access_km,
        destination_access_km = dest_info$access_km,
        origin_access_h = origin_access_h,
        destination_access_h = dest_info$access_time_h,
        transfer_penalty_h = TRANSFER_PENALTY_H,
        rail_total_time_h = NA_real_,
        geometry = st_sfc(st_geometrycollection(), crs = ANALYSIS_CRS)
      )
    )
  }

  dest_node_id <- dest_info$node_id
  path <- st_network_paths(
    rail_net,
    from = origin_node_id,
    to = dest_node_id,
    weights = "edge_time_h"
  )

  edge_ids <- unlist(path$edge_paths)

  if (!length(edge_ids)) {
    return(
      st_sf(
        end_icao = dest_code,
        rail_connected = FALSE,
        connector_feasible = TRUE,
        rail_km = NA_real_,
        high_speed_km = NA_real_,
        raw_high_speed_km = NA_real_,
        conventional_km = NA_real_,
        rail_in_vehicle_h = NA_real_,
        origin_access_km = origin_access_km,
        destination_access_km = dest_info$access_km,
        origin_access_h = origin_access_h,
        destination_access_h = dest_info$access_time_h,
        transfer_penalty_h = TRANSFER_PENALTY_H,
        rail_total_time_h = NA_real_,
        geometry = st_sfc(st_geometrycollection(), crs = ANALYSIS_CRS)
      )
    )
  }

  path_edges <- network_edges[edge_ids, ]
  path_edge_df <- path_edges %>%
    st_drop_geometry() %>%
    mutate(
      is_hsr = TYPE == "High speed",
      run_id = cumsum(coalesce(is_hsr != lag(is_hsr), TRUE))
    )

  hsr_runs <- path_edge_df %>%
    group_by(run_id, is_hsr) %>%
    summarise(run_km = sum(edge_length_km), .groups = "drop")

  effective_high_speed_km <- hsr_runs %>%
    filter(is_hsr, run_km >= MIN_CONTIGUOUS_HSR_KM) %>%
    summarise(value = sum(run_km), .groups = "drop") %>%
    pull(value)

  if (!length(effective_high_speed_km) || is.na(effective_high_speed_km)) {
    effective_high_speed_km <- 0
  }

  raw_high_speed_km <- path_edge_df %>%
    filter(is_hsr) %>%
    summarise(value = sum(edge_length_km), .groups = "drop") %>%
    pull(value)

  if (!length(raw_high_speed_km) || is.na(raw_high_speed_km)) {
    raw_high_speed_km <- 0
  }

  conventional_km <- sum(path_edge_df$edge_length_km) - effective_high_speed_km
  effective_in_vehicle_h <- (effective_high_speed_km / HSR_SPEED_KMH) +
    (conventional_km / CONVENTIONAL_SPEED_KMH)

  route_geom <- st_line_merge(st_union(st_geometry(path_edges)))

  st_sf(
    end_icao = dest_code,
    rail_connected = TRUE,
    connector_feasible = TRUE,
    rail_km = sum(path_edges$edge_length_km),
    high_speed_km = effective_high_speed_km,
    raw_high_speed_km = raw_high_speed_km,
    conventional_km = conventional_km,
    rail_in_vehicle_h = effective_in_vehicle_h,
    origin_access_km = origin_access_km,
    destination_access_km = dest_info$access_km,
    origin_access_h = origin_access_h,
    destination_access_h = dest_info$access_time_h,
    transfer_penalty_h = TRANSFER_PENALTY_H,
    rail_total_time_h = effective_in_vehicle_h + origin_access_h + dest_info$access_time_h + TRANSFER_PENALTY_H,
    geometry = st_sfc(route_geom, crs = ANALYSIS_CRS)
  )
}
```

```{r build-all-rail-routes}
rail_routes_sf <- do.call(
  rbind,
  lapply(barcelona_routes$end_icao, build_rail_route)
) %>%
  left_join(
    barcelona_routes %>%
      select(end_icao, end_airport, total_flights, route_passengers, raw_rows),
    by = "end_icao"
  ) %>%
  st_as_sf()

rail_route_summary <- rail_routes_sf %>%
  st_drop_geometry() %>%
  mutate(
    total_connector_km = origin_access_km + destination_access_km,
    total_connector_h = origin_access_h + destination_access_h,
    hsr_share = if_else(rail_connected & rail_km > 0, high_speed_km / rail_km, NA_real_),
    avg_kmh = if_else(rail_connected & rail_total_time_h > 0, rail_km / rail_total_time_h, NA_real_),
    connector_share = if_else(rail_connected & rail_total_time_h > 0, total_connector_h / rail_total_time_h, NA_real_)
  ) %>%
  arrange(rail_total_time_h)

rail_route_summary %>%
  select(
    end_airport,
    end_icao,
    rail_connected,
    rail_km,
    high_speed_km,
    conventional_km,
    rail_total_time_h
  ) %>%
  slice_head(n = 20)
```

```{r rail-routing-summary}
cat("Rail-connected Barcelona destinations:", sum(rail_routes_sf$rail_connected, na.rm = TRUE), "\n")
cat("Not rail-connected in this network:", sum(!rail_routes_sf$rail_connected, na.rm = TRUE), "\n")
cat(
  "Excluded by max airport-to-rail connector threshold:",
  sum(!rail_routes_sf$connector_feasible, na.rm = TRUE),
  "\n"
)
cat(
  "Within connector threshold but still disconnected in rail graph:",
  sum(rail_routes_sf$connector_feasible & !rail_routes_sf$rail_connected, na.rm = TRUE),
  "\n"
)

rail_routes_sf %>%
  st_drop_geometry() %>%
  summarise(
    median_time_h = median(rail_total_time_h, na.rm = TRUE),
    mean_time_h = mean(rail_total_time_h, na.rm = TRUE),
    max_time_h = max(rail_total_time_h, na.rm = TRUE)
  )
```

## Rail Diagnostics

These checks are here to test whether the travel-time surface looks plausible.
If almost every connected route uses a large share of high-speed track, or if
average end-to-end speeds cluster implausibly high, that is a sign the GL2017
typing or the network topology may still need refinement.

In the diagnostics below, `raw_high_speed_km` is the total routed length on
segments labelled `High speed`, while `high_speed_km` is the effective HSR
length that survives the contiguous-run threshold of `MIN_CONTIGUOUS_HSR_KM`.
Total rail time includes both airport connector times and a fixed transfer
penalty of `TRANSFER_PENALTY_H`.

```{r rail-diagnostics-summary}
rail_route_summary %>%
  filter(rail_connected) %>%
  summarise(
    connected_routes = n(),
    min_time_h = min(rail_total_time_h, na.rm = TRUE),
    q25_time_h = quantile(rail_total_time_h, 0.25, na.rm = TRUE),
    median_time_h = median(rail_total_time_h, na.rm = TRUE),
    mean_time_h = mean(rail_total_time_h, na.rm = TRUE),
    q75_time_h = quantile(rail_total_time_h, 0.75, na.rm = TRUE),
    max_time_h = max(rail_total_time_h, na.rm = TRUE),
    median_avg_kmh = median(avg_kmh, na.rm = TRUE),
    mean_avg_kmh = mean(avg_kmh, na.rm = TRUE),
    max_avg_kmh = max(avg_kmh, na.rm = TRUE)
  )
```

```{r routes-with-hsr}
routes_with_hsr <- rail_route_summary %>%
  filter(rail_connected, high_speed_km > 0) %>%
  select(
    end_airport,
    end_icao,
    rail_total_time_h,
    rail_km,
    raw_high_speed_km,
    high_speed_km,
    conventional_km,
    hsr_share,
    avg_kmh,
    route_passengers
  ) %>%
  arrange(desc(high_speed_km))

routes_with_hsr %>% slice_head(n = 40)
```

```{r routes-without-hsr}
routes_without_hsr <- rail_route_summary %>%
  filter(rail_connected, high_speed_km == 0) %>%
  select(
    end_airport,
    end_icao,
    rail_total_time_h,
    rail_km,
    raw_high_speed_km,
    total_connector_km,
    avg_kmh,
    route_passengers
  ) %>%
  arrange(rail_total_time_h)

routes_without_hsr
```

```{r suspicious-routes}
rail_route_summary %>%
  filter(rail_connected) %>%
  select(
    end_airport,
    end_icao,
    rail_total_time_h,
    rail_km,
    total_connector_km,
    total_connector_h,
    raw_high_speed_km,
    high_speed_km,
    conventional_km,
    hsr_share,
    avg_kmh,
    connector_share,
    route_passengers
  ) %>%
  arrange(desc(hsr_share), desc(avg_kmh), rail_total_time_h) %>%
  slice_head(n = 30)
```

## Presentation Figures

For presentation, the routed-path map is visually dense because many Barcelona
routes share the same rail corridors. The three figures below focus instead on:

- a simple origin-destination map from Barcelona
- the distribution of estimated rail times
- a ranked destination view

```{r presentation-data}
presentation_routes <- rail_route_summary %>%
  left_join(
    destination_airports %>%
      st_drop_geometry() %>%
      select(end_icao),
    by = "end_icao"
  ) %>%
  left_join(
    airports_clean %>%
      st_drop_geometry() %>%
      transmute(end_icao = ICAO, dest_lat = GeoPointLat, dest_lon = GeoPointLong),
    by = "end_icao"
  ) %>%
  left_join(
    destination_airports %>%
      mutate(dest_x = st_coordinates(.)[, 1], dest_y = st_coordinates(.)[, 2]) %>%
      st_drop_geometry() %>%
      select(end_icao, dest_x, dest_y),
    by = "end_icao"
  ) %>%
  mutate(
    time_bin = case_when(
      !connector_feasible ~ "Not rail-feasible",
      connector_feasible & !rail_connected ~ "Disconnected in network",
      rail_total_time_h < 4 ~ "< 4 h",
      rail_total_time_h < 6 ~ "4 to <6 h",
      rail_total_time_h < 8 ~ "6 to <8 h",
      TRUE ~ "8 h+"
    ),
    time_bin = factor(
      time_bin,
      levels = c("< 4 h", "4 to <6 h", "6 to <8 h", "8 h+", "Disconnected in network", "Not rail-feasible")
    )
  )

origin_coords <- st_coordinates(barcelona_airport)[1, 1:2]

straight_line_sfc <- st_sfc(
  lapply(seq_len(nrow(presentation_routes)), function(i) {
    st_linestring(
      matrix(
        c(
          origin_coords[1], origin_coords[2],
          presentation_routes$dest_x[i], presentation_routes$dest_y[i]
        ),
        ncol = 2,
        byrow = TRUE
      )
    )
  }),
  crs = ANALYSIS_CRS
)

presentation_lines <- st_sf(
  presentation_routes %>%
    select(end_icao, end_airport, total_flights, route_passengers, rail_total_time_h, time_bin),
  geometry = straight_line_sfc
)

presentation_points <- destination_airports %>%
  left_join(
    presentation_routes %>%
      select(end_icao, end_airport, total_flights, route_passengers, rail_total_time_h, time_bin),
    by = "end_icao"
  )

time_bin_colors <- c(
  "< 4 h" = "#1b9e77",
  "4 to <6 h" = "#66a61e",
  "6 to <8 h" = "#e6ab02",
  "8 h+" = "#d95f02",
  "Disconnected in network" = "#7570b3",
  "Not rail-feasible" = "#bdbdbd"
)
```

### BCN Route and Passenger Summary

```{r bcn-summary}
bcn_summary <- presentation_routes %>%
  st_drop_geometry() %>%
  mutate(
    counted_passengers = coalesce(route_passengers, 0)
  ) %>%
  group_by(time_bin) %>%
  summarise(
    routes = n(),
    passengers = sum(counted_passengers, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    route_pct = routes / sum(routes),
    passenger_pct = passengers / sum(passengers)
  )

bcn_summary
```

```{r bcn-summary-collapsed}
bcn_summary_collapsed <- presentation_routes %>%
  st_drop_geometry() %>%
  mutate(
    counted_passengers = coalesce(route_passengers, 0),
    rail_substitutable = case_when(
      rail_connected & rail_total_time_h < 6 ~ "Rail-competitive (<6 h)",
      rail_connected ~ "Rail-connected but slower (6 h+)",
      TRUE ~ "Not rail-feasible / disconnected"
    )
  ) %>%
  group_by(rail_substitutable) %>%
  summarise(
    routes = n(),
    passengers = sum(counted_passengers, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    route_pct = routes / sum(routes),
    passenger_pct = passengers / sum(passengers)
  )

bcn_summary_collapsed
```

## HSR Alternative Flags

For the BCN-focused notebook, I flag routes as plausible high-speed-rail
alternatives when they are:

- connector-feasible
- rail-connected in the GL2017 passenger network
- below a simple rail-time threshold of 6 hours

This is a descriptive substitution flag, not yet a full generalized-cost test.

```{r hsr-alternative-flags}
hsr_flag_routes <- presentation_routes %>%
  st_drop_geometry() %>%
  mutate(
    hsr_alternative = rail_connected & rail_total_time_h < 6,
    counted_passengers = coalesce(route_passengers, 0),
    counted_flights = coalesce(total_flights, 0)
  )

hsr_flag_routes %>%
  select(
    end_airport,
    end_icao,
    total_flights,
    route_passengers,
    rail_total_time_h,
    hsr_alternative
  ) %>%
  arrange(rail_total_time_h)
```

```{r hsr-alternative-summary}
hsr_alternative_summary <- hsr_flag_routes %>%
  summarise(
    total_routes = n(),
    total_flights = sum(counted_flights, na.rm = TRUE),
    total_passengers = sum(counted_passengers, na.rm = TRUE),
    alternative_routes = sum(hsr_alternative, na.rm = TRUE),
    alternative_route_share = alternative_routes / total_routes,
    alternative_flights = sum(counted_flights[hsr_alternative], na.rm = TRUE),
    alternative_flight_share = alternative_flights / total_flights,
    alternative_passengers = sum(counted_passengers[hsr_alternative], na.rm = TRUE),
    alternative_passenger_share = alternative_passengers / total_passengers
  )

hsr_alternative_summary
```

```{r hsr-alternative-by-bin}
hsr_flag_routes %>%
  mutate(
    alternative_group = case_when(
      hsr_alternative ~ "Flagged as HSR alternative",
      rail_connected ~ "Rail-connected but slower (6 h+)",
      TRUE ~ "Not rail-feasible / disconnected"
    )
  ) %>%
  group_by(alternative_group) %>%
  summarise(
    routes = n(),
    flights = sum(counted_flights, na.rm = TRUE),
    passengers = sum(counted_passengers, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    route_pct = routes / sum(routes),
    flight_pct = flights / sum(flights),
    passenger_pct = passengers / sum(passengers)
  )
```

### Figure 1: Straight-Line Map from Barcelona

```{r od-time-map, fig.cap="Barcelona origin-destination map coloured by estimated rail travel-time class"}
od_time_map <- ggplot() +
  geom_sf(data = countries, fill = "#F8F8F6", color = "#D0D5D9", linewidth = 0.16) +
  geom_sf(
    data = presentation_lines,
    aes(color = time_bin),
    linewidth = 0.45,
    alpha = 0.55,
    lineend = "round",
    show.legend = "line"
  ) +
  geom_sf(
    data = presentation_points,
    aes(fill = time_bin),
    shape = 21,
    size = 2.2,
    color = "#1E2D3B",
    stroke = 0.25
  ) +
  geom_sf(
    data = barcelona_airport,
    shape = 21,
    size = 5,
    fill = "#F25F5C",
    color = "#5C1A18",
    stroke = 0.8
  ) +
  geom_text_repel(
    data = destination_airports %>%
      left_join(rail_routes_sf %>% st_drop_geometry(), by = "end_icao") %>%
      filter(rail_connected) %>%
      slice_min(order_by = rail_total_time_h, n = 12) %>%
      cbind(st_coordinates(.)),
    aes(X, Y, label = end_airport),
    size = 3.1,
    box.padding = 0.35,
    point.padding = 0.2,
    min.segment.length = 0
  ) +
  scale_color_manual(values = time_bin_colors, name = NULL, drop = FALSE) +
  scale_fill_manual(values = time_bin_colors, guide = "none", drop = FALSE) +
  coord_sf(xlim = map_xlim, ylim = map_ylim, expand = FALSE) +
  labs(
    title = "Barcelona-to-Destination Rail Time Classes",
    subtitle = paste0(
      "Straight origin-destination lines for readability; classes reflect routed rail-time estimates with a ",
      MAX_CONNECTOR_KM,
      " km snap cap"
    ),
    caption = "Source files: data/raw/shapefiles/railways_GL2017_EU.shp and data/clean/all_airports.gpkg"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.background = element_rect(fill = "#EEF4F7", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "#D9E3E9", linewidth = 0.22),
    panel.grid.minor = element_blank(),
    axis.title = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )

od_time_map
```

```{r save-od-time-map, eval=FALSE}
ggsave(
  file.path(outputs, "barcelona_el_prat_od_time_bins.png"),
  od_time_map,
  width = 13,
  height = 8,
  dpi = 300,
  bg = "white"
)
```

### Figure 2: Histogram of Rail Times

```{r rail-time-histogram, fig.cap="Distribution of Barcelona rail travel times to western-European destinations"}
rail_time_hist <- presentation_routes %>%
  filter(rail_connected) %>%
  ggplot(aes(x = rail_total_time_h)) +
  geom_histogram(binwidth = 1, boundary = 0, fill = "#2C4C7E", color = "white", alpha = 0.9) +
  geom_vline(xintercept = median(presentation_routes$rail_total_time_h[presentation_routes$rail_connected], na.rm = TRUE),
             color = "#D95F02", linewidth = 0.9, linetype = "dashed") +
  labs(
    title = "Distribution of Estimated Rail Travel Times from Barcelona",
    subtitle = "Only destinations that are connector-feasible and rail-connected are included",
    x = "Estimated rail travel time (hours)",
    y = "Number of destinations",
    caption = "Dashed line marks the median connected-destination rail time"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

rail_time_hist
```

```{r save-rail-time-histogram, eval=FALSE}
ggsave(
  file.path(outputs, "barcelona_el_prat_rail_time_histogram.png"),
  rail_time_hist,
  width = 10,
  height = 6,
  dpi = 300,
  bg = "white"
)
```

### Figure 3: Ranked Destination Rail Times

```{r ranked-rail-times, fig.cap="Ranked Barcelona destinations by estimated rail travel time"}
ranked_destinations <- presentation_routes %>%
  filter(rail_connected) %>%
  arrange(rail_total_time_h) %>%
  mutate(end_airport = factor(end_airport, levels = rev(end_airport))) %>%
  slice_head(n = 30)

ranked_rail_times <- ggplot(ranked_destinations, aes(x = rail_total_time_h, y = end_airport, color = time_bin)) +
  geom_segment(
    aes(x = 0, xend = rail_total_time_h, y = end_airport, yend = end_airport),
    linewidth = 0.7,
    alpha = 0.7
  ) +
  geom_point(aes(size = route_passengers), alpha = 0.9) +
  scale_color_manual(values = time_bin_colors, name = NULL, drop = FALSE) +
  scale_size_continuous(range = c(2, 6), labels = label_number(big.mark = ","), name = "Route passengers") +
  labs(
    title = "Fastest Rail-Connected Barcelona Destinations",
    subtitle = "Top 30 destinations ranked by estimated rail travel time",
    x = "Estimated rail travel time (hours)",
    y = NULL,
    caption = "Point size reflects route passengers"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

ranked_rail_times
```

```{r save-ranked-rail-times, eval=FALSE}
ggsave(
  file.path(outputs, "barcelona_el_prat_ranked_rail_times.png"),
  ranked_rail_times,
  width = 10,
  height = 9,
  dpi = 300,
  bg = "white"
)
```

## Export a Route Summary Table

```{r export-table, eval=FALSE}
rail_routes_sf %>%
  st_drop_geometry() %>%
  arrange(rail_total_time_h) %>%
  write_csv(file.path(outputs, "barcelona_el_prat_rail_route_summary.csv"))
```

## Notes

- `master_routes_clean.csv` contains duplicate Barcelona destination rows for a
  small number of ICAOs, so the notebook collapses to one row per `end_icao`.
- `all_airports.gpkg` contains duplicate airport ICAOs for `EDDB` and `EDLW`;
  the notebook resolves those by preferring non-station entries and then longer
  airport names.
- Islands and disconnected segments in the rail network correctly return no
  rail path.
