---
title: "Short-Haul Flight Substitution by High-Speed Rail in Europe"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

This notebook supplemented my EDA for my final project in my
[Geospatial Data Science and Economic Spatial Models](https://brunoconteleite.github.io/08-geospatial-ds-bse/)
class for the Spring 2026 Term. After my EDA, my goal was to present a
preliminary analysis to determine if this project is reasonable for the final
capstone.

**Research question:** Which short-haul European flight routes are substitutable
by rail based on estimated travel times, and what policy levers make that
substitution welfare-improving?

**Approach:**

- Spatial analysis of airport access to the European rail network
- Flight time and rail time modelling to identify substitutable routes
- Carbon-adjusted market access framework (Allen & Arkolakis, 2014) to measure
  connectivity change
- Sensitivity analysis across carbon prices and airport overhead assumptions

**Data:** Natural Earth rail, GL2017 rail, OpenFlights (airports + routes),
cleaned Eurostat airport passenger totals already stored in this repo

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 14,
  fig.height = 8.75
)
```

```{r libraries}
library(sf)
library(prettymapr)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggspatial)   # annotation_map_tile() — replaces contextily
library(ggrepel)
library(patchwork)
library(scales)
```

```{r paths}
# here::here() finds the project root via .Rproj regardless of working directory.
# This is necessary in Rmd: knitr sets WD to the Rmd's folder (scripts/), not the
# project root, so the commandArgs() approach used in plain .R scripts won't work.
library(here)
source(here("scripts", "paths.R"))

dir.create(file.path(root_dir, "slides"), showWarnings = FALSE)

# Plot constants (mirror Python's W/H/DPI/colour scheme)
C_HSR    <- "#e63946"
C_CONV   <- "#cccccc"
C_APHSR  <- "#2176ae"
C_APNONE <- "#aaaaaa"
SOURCE   <- "Data: Natural Earth rail, GL2017 rail, OpenFlights, cleaned Eurostat airport totals"
ANALYSIS_CRS <- 3035
RAIL_SPEED <- 800
LOCAL_ACCESS_SPEED <- 60
STATION_PENALTY <- 0.75

# Minimal theme for all maps
theme_map <- function(...) {
  theme_void(...) +
    theme(
      plot.title    = element_text(face = "bold", size = 16, hjust = 0),
      plot.subtitle = element_text(size = 10, hjust = 0, color = "#444444"),
      plot.caption  = element_text(size = 7.5, color = "#999999",
                                   face = "italic", hjust = 1),
      legend.position = "bottom"
    )
}
```

---

## 1. European Rail Infrastructure

The original workflow depended on a GL2017 rail file with a high-speed flag.
That is useful for visualization, but it is not reliable enough to define
substitutability in the model. I now use the more complete Natural Earth rail
network for calculations and define rail feasibility functionally using travel
time rather than infrastructure labels.

GL2017 is still kept as an optional visual overlay so I can compare modeled
rail-feasible routes against where an explicit high-speed label exists. It is
not used in the calculations below.

```{r load-railways}
countries <- st_read(file.path(data_clean, "ne_10m_countries.gpkg"), quiet = TRUE) %>%
  st_make_valid() %>%
  st_transform(ANALYSIS_CRS)

countries_mask <- st_union(countries)

rail_ne <- st_read(
  paste0(
    "/vsizip/",
    normalizePath(
      file.path(data_clean, "shapefiles", "ne_10m_railroads.zip"),
      winslash = "/", mustWork = TRUE
    )
  ),
  quiet = TRUE
) %>%
  st_make_valid() %>%
  st_transform(ANALYSIS_CRS) %>%
  st_intersection(countries_mask)

rail_gl2017 <- st_read(
  file.path(data_clean, "shapefiles", "railways_GL2017_EU.shp"),
  quiet = TRUE
) %>%
  st_make_valid() %>%
  st_transform(ANALYSIS_CRS) %>%
  st_intersection(countries_mask)

gl2017_hs <- rail_gl2017 %>% filter(TYPE == "High speed")

cat(sprintf("Natural Earth rail segments: %s\n", format(nrow(rail_ne), big.mark = ",")))
cat(sprintf("GL2017 rail segments:        %s\n", format(nrow(rail_gl2017), big.mark = ",")))
cat(sprintf("GL2017 'High speed':         %s\n", format(nrow(gl2017_hs), big.mark = ",")))
cat(sprintf("Analysis CRS:                %s\n", st_crs(rail_ne)$input))

cat("\nNatural Earth feature classes:\n")
print(rail_ne %>% st_drop_geometry() %>% count(featurecla, sort = TRUE))
```

```{r plot-rail, fig.cap="European rail network — conventional and high speed"}
p_rail <- ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
  geom_sf(data = rail_ne, color = "forestgreen", linewidth = 0.3, alpha = 0.65) +
  geom_sf(data = gl2017_hs, color = C_HSR, linewidth = 0.5, alpha = 0.8) +
  labs(
    title   = "European Rail Network for the Accessibility Model",
    subtitle = "Natural Earth rail is the modeling base; GL2017 high-speed labels are shown only as an optional overlay",
    caption = SOURCE
  ) +
  theme_map()

ggsave(file.path(root_dir, "slides", "01_hsr_network.png"),
       p_rail, width = 14, height = 8.75, dpi = 150, bg = "white")
p_rail
```

---

## 2. Airport and Flight Network Data

Airport and route data is from OpenFlights. I filtered to a European bounding
box, kept only airports with an IATA code (which excludes military and private
airfields), and dropped airports that don't appear in the routes file.

**Filtering decisions:**

- Bounding box: lat 34–72 N, lon 25 W–45 E
- IATA code required (civilian/commercial airports only)
- Must appear in `routes.txt`
- Routes filtered to Europe–Europe pairs only

```{r load-airports}
ap_cols <- c("id", "name", "city", "country", "iata", "icao",
             "lat", "lon", "alt", "tz", "dst", "tz_db", "type", "source")
airports_raw <- read.csv(
  file.path(data_clean, "openflights", "airports.txt"),
  header = FALSE, col.names = ap_cols,
  na.strings = "\\N", stringsAsFactors = FALSE
)

rt_cols <- c("airline", "airline_id", "src_iata", "src_id",
             "dst_iata", "dst_id", "codeshare", "stops", "equipment")
routes_raw <- read.csv(
  file.path(data_clean, "openflights", "routes.txt"),
  header = FALSE, col.names = rt_cols,
  na.strings = "\\N", stringsAsFactors = FALSE
)
routes_raw <- routes_raw %>%
  mutate(
    src_id = suppressWarnings(as.integer(src_id)),
    dst_id = suppressWarnings(as.integer(dst_id))
  ) %>%
  filter(!is.na(src_id), !is.na(dst_id))

# Filter to civil European airports with IATA codes
europe_civil <- airports_raw %>%
  filter(
    between(lat, 34, 72),
    between(lon, -25, 45),
    type == "airport",
    !is.na(iata)
  )

eu_ids    <- unique(europe_civil$id)
eu_routes <- routes_raw %>%
  filter(src_id %in% eu_ids, dst_id %in% eu_ids)

# Keep only airports that appear in the routes file
eu_ids       <- union(eu_routes$src_id, eu_routes$dst_id)
europe_civil <- europe_civil %>% filter(id %in% eu_ids)

# Attach lat/lon to routes
ap_coords <- europe_civil %>% select(id, lat, lon)
eu_routes <- eu_routes %>%
  left_join(ap_coords %>% rename(src_id = id, src_lat = lat, src_lon = lon),
            by = "src_id") %>%
  left_join(ap_coords %>% rename(dst_id = id, dst_lat = lat, dst_lon = lon),
            by = "dst_id")

gdf <- europe_civil %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(3857)

cat(sprintf("Civil European airports: %s\n", format(nrow(europe_civil), big.mark = ",")))
cat(sprintf("Europe–Europe routes:    %s\n", format(nrow(eu_routes),    big.mark = ",")))
```

```{r plot-flight-network}
routes_sf <- eu_routes %>%
  rowwise() %>%
  mutate(geometry = st_sfc(
    st_linestring(matrix(c(src_lon, src_lat, dst_lon, dst_lat), ncol = 2, byrow = TRUE)),
    crs = 4326
  )) %>%
  ungroup() %>%
  st_as_sf() %>%
  st_transform(3857)

degree_all <- c(eu_routes$src_id, eu_routes$dst_id) %>%
  table() %>% as.data.frame() %>%
  setNames(c("id", "degree")) %>%
  mutate(id = as.integer(as.character(id)))

gdf_sized <- gdf %>%
  left_join(degree_all, by = "id") %>%
  filter(!is.na(degree))

p_network <- ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
  geom_sf(data = routes_sf,  color = C_APHSR, linewidth = 0.15, alpha = 0.10) +
  geom_sf(data = gdf_sized,  color = C_HSR,   aes(size = log1p(degree)),
          alpha = 0.7, show.legend = FALSE) +
  scale_size_continuous(range = c(0.5, 4)) +
  labs(
    title   = sprintf("European Flight Network  (%s routes  |  %s airports)",
                      format(nrow(routes_sf), big.mark = ","),
                      format(nrow(gdf_sized), big.mark = ",")),
    caption = SOURCE
  ) +
  theme_map()

ggsave(file.path(root_dir, "slides", "02_eu_flight_network.png"),
       p_network, width = 14, height = 8.75, dpi = 150, bg = "white")
p_network
```

That's overwhelming — later we will filter out for short flights that we
actually care about, as well as passenger counts to focus on high-volume travel
routes.

---

## 3. Passenger Demand — Eurostat `avia_paoa`

Passenger data comes from the cleaned airport passenger layer already stored in
this repo. That keeps the notebook reproducible for classmates without relying
on a live API call during rendering. Airports below the reporting threshold
(mostly small regional ones) are kept but excluded from demand-weighted
analysis.

**Notes:**

- Ideally I'd have O-D data, but total airport volume is what's available here
- Gravity score = origin passengers × destination passengers — a rough but
  standard demand proxy

```{r load-pax}
pax_clean <- st_read(
  file.path(data_clean, "airports_pax_2019.gpkg"),
  quiet = TRUE
) %>%
  st_drop_geometry() %>%
  transmute(
    icao = ICAO,
    passengers = airport_total_passengers,
    year = year
  ) %>%
  filter(!is.na(icao), !is.na(passengers)) %>%
  distinct(icao, .keep_all = TRUE)

latest_year <- pax_clean %>%
  summarise(year = max(year, na.rm = TRUE)) %>%
  pull(year)

europe_civil_pax <- europe_civil %>%
  left_join(pax_clean %>% select(icao, passengers), by = "icao")

matched <- sum(!is.na(europe_civil_pax$passengers))
cat(sprintf("Year: %d  |  Airports matched to cleaned passenger totals: %d / %d\n",
            latest_year, matched, nrow(europe_civil_pax)))
cat("\nTop 10 airports by passengers:\n")
europe_civil_pax %>%
  filter(!is.na(passengers)) %>%
  arrange(desc(passengers)) %>%
  slice_head(n = 10) %>%
  select(name, iata, passengers) %>%
  print()
```

```{r plot-pax}
gdf_pax <- europe_civil_pax %>%
  distinct(id, .keep_all = TRUE) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(3857)

has_d <- gdf_pax %>% filter(!is.na(passengers))
no_d  <- gdf_pax %>% filter( is.na(passengers))

pax_norm  <- (has_d$passengers - min(has_d$passengers, na.rm = TRUE)) /
             (max(has_d$passengers, na.rm = TRUE) - min(has_d$passengers, na.rm = TRUE))
has_d$sz  <- 3 + 10 * sqrt(pax_norm)

top10_airports <- has_d %>% slice_max(passengers, n = 10)

p_pax <- ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
  geom_sf(data = no_d,  color = C_APNONE, size = 1.5, alpha = 0.4) +
  geom_sf(data = has_d, color = C_APHSR,  aes(size = sz), alpha = 0.75,
          show.legend = FALSE) +
  geom_text_repel(
    data  = top10_airports,
    aes(label = iata, geometry = geometry),
    stat  = "sf_coordinates",
    size  = 2.5, fontface = "bold", color = "#222222",
    min.segment.length = 0
  ) +
  scale_size_identity() +
  labs(
    title   = sprintf("European Civil Airports  -  Annual Passenger Traffic  (%d)",
                      latest_year),
    caption = SOURCE
  ) +
  theme_map()

ggsave(file.path(root_dir, "slides", "03_airport_passengers.png"),
       p_pax, width = 14, height = 8.75, dpi = 150, bg = "white")
p_pax
```

---

## 4. Identifying Short-Haul Routes

Flight time is estimated using a simplified model:
`dist_km / 800 km/h + 0.75 h overhead`, where 800 km/h approximates the ground
speed of a narrow-body aircraft (A320/737) and 0.75 h captures boarding, taxi,
and arrival. Routes with estimated flight time ≤ 2.5 h are classified as
short-haul.

Great-circle (Haversine) distance is used as a conservative proxy — actual
flight paths may be longer, making this a lower bound on true flight time.

**Notes:**

- 0.75 h is a conservative lower bound for airport overhead — sensitivity is
  tested later.
- Could pull actual flight time logs later. Most research suggests that
  door-to-door time is most important in analysing rail vs air trade-offs.

```{r haversine}
haversine_km <- function(lat1, lon1, lat2, lon2) {
  R    <- 6371
  dlat <- (lat2 - lat1) * pi / 180
  dlon <- (lon2 - lon1) * pi / 180
  a    <- sin(dlat / 2)^2 +
          cos(lat1 * pi / 180) * cos(lat2 * pi / 180) * sin(dlon / 2)^2
  2 * R * asin(sqrt(a))
}

eu_routes <- eu_routes %>%
  mutate(
    dist_km       = haversine_km(src_lat, src_lon, dst_lat, dst_lon),
    flight_time_h = dist_km / 800 + 0.75
  )

short <- eu_routes %>% filter(flight_time_h <= 2.5)

cat(sprintf("Total EU routes:        %s\n",  format(nrow(eu_routes), big.mark = ",")))
cat(sprintf("Short-haul (≤ 2.5 h):  %s  (%.1f%%)\n",
            format(nrow(short), big.mark = ","),
            nrow(short) / nrow(eu_routes) * 100))
cat(sprintf("Distance range:         %.0f – %.0f km\n",
            min(short$dist_km), max(short$dist_km)))
```

Plot these routes, coloured by flight time:

```{r plot-short-haul}
short_sf <- short %>%
  rowwise() %>%
  mutate(geometry = st_sfc(
    st_linestring(matrix(c(src_lon, src_lat, dst_lon, dst_lat), ncol = 2, byrow = TRUE)),
    crs = 4326
  )) %>%
  ungroup() %>%
  st_as_sf() %>%
  st_transform(3857)

degree_s <- c(short$src_id, short$dst_id) %>%
  table() %>% as.data.frame() %>%
  setNames(c("id", "degree")) %>%
  mutate(id = as.integer(as.character(id)))

gdf_civil <- europe_civil %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(3857)

gdf_s <- gdf_civil %>%
  inner_join(degree_s, by = "id")

p_short <- ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
  geom_sf(data = short_sf, aes(color = flight_time_h),
          linewidth = 0.4, alpha = 0.5) +
  geom_sf(data = gdf_s, color = "#333333",
          aes(size = log1p(degree)), alpha = 0.6, show.legend = FALSE) +
  scale_color_gradient2(
    low = "#1a9850", mid = "#ffffbf", high = "#d73027",
    midpoint = median(short_sf$flight_time_h),
    name = "Estimated flight time (hours)",
    guide = guide_colorbar(title.position = "top", barwidth = 15)
  ) +
  scale_size_continuous(range = c(0.5, 3.5)) +
  labs(
    title    = sprintf("Short-Haul Flight Routes  (≤ 2.5 h  |  n = %s routes)",
                       format(nrow(short), big.mark = ",")),
    subtitle = "Green = short hop   |   Red = approaching 2.5 h",
    caption  = SOURCE
  ) +
  theme_map()

ggsave(file.path(root_dir, "slides", "04_short_haul_routes.png"),
       p_short, width = 14, height = 8.75, dpi = 150, bg = "white")
p_short
```

Again, this is also pretty messy. We do see some shorter flights from the UK,
as well as Istanbul, others in northern Europe as well.

---

## Refactored accessibility model

The sections below implement the time-based rail accessibility model that
replaces the older HSR-buffer workflow. They are the chunks that should be used
for interpretation and any downstream calculations.

```{r rail-access}
gdf_civil <- europe_civil %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(ANALYSIS_CRS)

nearest_rail_idx <- st_nearest_feature(gdf_civil, rail_ne)
nearest_rail_dist_m <- st_distance(
  gdf_civil,
  rail_ne[nearest_rail_idx, ],
  by_element = TRUE
)

airport_access <- gdf_civil %>%
  mutate(
    dist_to_rail_km = as.numeric(nearest_rail_dist_m) / 1000,
    access_time_h = dist_to_rail_km / LOCAL_ACCESS_SPEED
  )

airport_access_lookup <- airport_access %>%
  st_drop_geometry() %>%
  select(id, iata, dist_to_rail_km, access_time_h)
```

```{r plot-rail-access}
p_buf <- ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
  geom_sf(data = rail_ne, color = "#888888", linewidth = 0.35, alpha = 0.55) +
  geom_sf(data = airport_access, aes(color = dist_to_rail_km), size = 2.2, alpha = 0.85) +
  scale_color_gradientn(
    colours = c("#1a9850", "#91cf60", "#fee08b", "#fc8d59", "#d73027"),
    name = "Distance to nearest rail (km)",
    guide = guide_colorbar(title.position = "top", barwidth = 15)
  ) +
  labs(
    title = "Airport-to-Rail Accessibility",
    subtitle = "Each airport is linked to the nearest Natural Earth rail segment; rail feasibility is defined using GIS distance, not HSR labels",
    caption = SOURCE
  ) +
  theme_map()

ggsave(file.path(root_dir, "slides", "05_buffer_methodology.png"),
       p_buf, width = 14, height = 8.75, dpi = 150, bg = "white")
p_buf
```

```{r plot-rail-access-mismatch}
airport_access_mismatch <- airport_access %>%
  mutate(
    gl2017_hs_near = lengths(st_intersects(geometry, st_buffer(gl2017_hs, 5000))) > 0,
    mismatch = case_when(
      access_time_h <= 0.25 & !gl2017_hs_near ~ "Rail-feasible, no GL2017 HSR nearby",
      access_time_h <= 0.25 &  gl2017_hs_near ~ "Rail-feasible, GL2017 HSR nearby",
      TRUE ~ "Longer local access"
    )
  )

p_flood <- ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
  geom_sf(data = rail_ne, color = "#bdbdbd", linewidth = 0.25, alpha = 0.45) +
  geom_sf(data = gl2017_hs, color = C_HSR, linewidth = 0.45, alpha = 0.55) +
  geom_sf(data = airport_access_mismatch, aes(color = mismatch), size = 2.3, alpha = 0.85) +
  scale_color_manual(
    values = c(
      "Rail-feasible, no GL2017 HSR nearby" = "#f4a261",
      "Rail-feasible, GL2017 HSR nearby" = C_APHSR,
      "Longer local access" = C_APNONE
    ),
    name = NULL
  ) +
  labs(
    title = "Modeled Rail Accessibility vs GL2017 Overlay",
    subtitle = "Optional comparison: where the time-based model finds rail-feasible access without a nearby GL2017 high-speed segment",
    caption = SOURCE
  ) +
  theme_map()

ggsave(file.path(root_dir, "slides", "06_hsr_flood_fill.png"),
       p_flood, width = 14, height = 8.75, dpi = 150, bg = "white")
p_flood
```

```{r substitutable-routes-refactor}
pax_lookup <- europe_civil_pax %>%
  distinct(id, .keep_all = TRUE) %>%
  filter(!is.na(passengers)) %>%
  select(id, name, passengers)

routes_time <- eu_routes %>%
  left_join(
    airport_access_lookup %>%
      rename(src_id = id, src_dist_to_rail_km = dist_to_rail_km, src_access_time_h = access_time_h),
    by = "src_id"
  ) %>%
  left_join(
    airport_access_lookup %>%
      rename(dst_id = id, dst_dist_to_rail_km = dist_to_rail_km, dst_access_time_h = access_time_h),
    by = "dst_id"
  ) %>%
  mutate(
    rail_time_h = dist_km / RAIL_SPEED + src_access_time_h + dst_access_time_h + STATION_PENALTY,
    substitutable = rail_time_h <= 2.5
  ) %>%
  left_join(pax_lookup %>% rename(src_id = id, src_name = name, src_pax = passengers), by = "src_id") %>%
  left_join(pax_lookup %>% rename(dst_id = id, dst_name = name, dst_pax = passengers), by = "dst_id") %>%
  mutate(
    gravity = src_pax * dst_pax,
    route_key = map2_chr(src_id, dst_id, ~ paste(sort(c(.x, .y)), collapse = "_"))
  )

sub_routes <- routes_time %>%
  filter(substitutable) %>%
  arrange(desc(gravity)) %>%
  distinct(route_key, .keep_all = TRUE)
```


## 5. Spatial Analysis — HSR Airport Access

I defined airport HSR access using a 5 km buffer around each airport,
intersected with the HSR layer. From there I ran a flood-fill: at each step
the accessible HSR network is buffered by 200 m and any touching segments get
added. Converged at iteration 13.

The flood-fill matters because an airport near one segment effectively gets
access to the whole connected corridor, not just that one piece.

**Notes:**

- LHR: 0 substitutable routes — no direct HSR connection
- CDG: 21 substitutable routes — integrated TGV terminal
- 5 km cutoff is arbitrary — ideally I'd also check for transit connections to
  HSR stations beyond the buffer
- Focused on HSR line edges rather than stations — a station-level dataset
  would be more precise

```{r flood-fill, eval=FALSE, include=FALSE}
airport_buf <- gdf_civil %>%
  st_transform(3857) %>%          # hs is in 3857
  st_buffer(5000)

civil_buf_union <- st_union(airport_buf)
hs_union        <- st_union(hs)

# Flood-fill: expand to the full connected HSR network
accessible <- as.logical(st_intersects(hs, civil_buf_union, sparse = FALSE))

for (i in seq_len(100)) {
  prev     <- sum(accessible)
  expanded <- hs$geometry[accessible] %>% st_union() %>% st_buffer(200)
  accessible <- as.logical(st_intersects(hs, expanded, sparse = FALSE))
  if (sum(accessible) == prev) {
    cat(sprintf("Converged at iteration %d\n", i))
    break
  }
}

hs_accessible     <- hs[ accessible, ]
hs_not_accessible <- hs[!accessible, ]
civil_near_hsr    <- gdf_civil %>%
  filter(as.logical(st_intersects(st_transform(., 3857) %>% st_buffer(5000),
                                  hs_union, sparse = FALSE)))
civil_no_hsr      <- gdf_civil %>% filter(!id %in% civil_near_hsr$id)

cat(sprintf("HSR segments accessible from airports: %d / %d\n",
            nrow(hs_accessible), nrow(hs)))
cat(sprintf("Civil airports with HSR access:        %d / %d\n",
            nrow(civil_near_hsr), nrow(gdf_civil)))
```

```{r plot-buffer, eval=FALSE, include=FALSE}
hs_near       <- hs %>% filter(as.logical(st_intersects(., civil_buf_union, sparse = FALSE)))
bufs_near_hsr <- airport_buf %>%
  filter(as.logical(st_intersects(., hs_union, sparse = FALSE)))

p_buf <- ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
  geom_sf(data = hs,            color = "#888888", linewidth = 0.6, alpha = 0.8) +
  geom_sf(data = hs_near,       color = C_HSR,     linewidth = 1.2) +
  geom_sf(data = bufs_near_hsr, fill  = C_APHSR,   color = C_APHSR,
          alpha = 0.25, linewidth = 0.5) +
  labs(
    title    = "Methodology: 5 km Airport Buffer  ∩  HSR Network",
    subtitle = "Identifies HSR segments directly accessible from airport terminals",
    caption  = SOURCE
  ) +
  theme_map()

ggsave(file.path(root_dir, "slides", "05_buffer_methodology.png"),
       p_buf, width = 14, height = 8.75, dpi = 150, bg = "white")
p_buf
```

```{r plot-flood-fill, eval=FALSE, include=FALSE}
p_flood <- ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
  geom_sf(data = hs_not_accessible, color = "#f4a261", linewidth = 0.6) +
  geom_sf(data = hs_accessible,     color = C_HSR,     linewidth = 0.9) +
  geom_sf(data = st_transform(civil_no_hsr,   3857), color = C_APNONE,
          size = 1.5, alpha = 0.5) +
  geom_sf(data = st_transform(civil_near_hsr, 3857), color = C_APHSR,
          size = 3,   alpha = 0.85) +
  labs(
    title    = "HSR Network Reachable from Civil Airports (flood-fill from 5 km catchment)",
    subtitle = "Airports trigger access to entire geometrically-connected HSR sub-network",
    caption  = SOURCE
  ) +
  theme_map()

ggsave(file.path(root_dir, "slides", "06_hsr_flood_fill.png"),
       p_flood, width = 14, height = 8.75, dpi = 150, bg = "white")
p_flood
```

---

## 6. Substitutable Routes and Demand Proxy

A route is substitutable if: (1) flight time is under 2.5 h, and (2) both
endpoints are within 5 km of HSR. Demand is proxied by a gravity score —
origin passengers × destination passengers — which is standard in bilateral
transport models.

**Notes:**

- Deduplicated so A–B and B–A count as one route
- Airports not matched to Eurostat are excluded from the ranked list

```{r substitutable-routes, eval=FALSE, include=FALSE}
hsr_ids  <- civil_near_hsr$id

pax_lookup <- europe_civil_pax %>%
  distinct(id, .keep_all = TRUE) %>%
  filter(!is.na(passengers)) %>%
  select(id, name, passengers)

both_hsr <- short %>%
  filter(src_id %in% hsr_ids, dst_id %in% hsr_ids) %>%
  left_join(pax_lookup %>% rename(src_id = id, src_name = name, src_pax = passengers),
            by = "src_id") %>%
  left_join(pax_lookup %>% rename(dst_id = id, dst_name = name, dst_pax = passengers),
            by = "dst_id") %>%
  mutate(
    gravity   = src_pax * dst_pax,
    route_key = map2_chr(src_id, dst_id,
                         ~ paste(sort(c(.x, .y)), collapse = "_"))
  )

both_hsr_dedup <- both_hsr %>%
  arrange(desc(gravity)) %>%
  distinct(route_key, .keep_all = TRUE)

cat(sprintf("Short-haul routes total:            %s\n",  format(nrow(short),           big.mark = ",")))
cat(sprintf("Both endpoints have HSR access:     %s\n",  format(nrow(both_hsr_dedup),  big.mark = ",")))
cat(sprintf("With full passenger data:           %s\n",  format(sum(!is.na(both_hsr_dedup$gravity)), big.mark = ",")))

cat("\nTop 15 by gravity score:\n")
both_hsr_dedup %>%
  filter(!is.na(gravity)) %>%
  slice_max(gravity, n = 15) %>%
  transmute(
    src_iata, src_name, dst_iata, dst_name,
    dist_km       = round(dist_km),
    flight_time_h = round(flight_time_h, 2),
    gravity       = paste0(round(gravity / 1e9, 2), "B")
  ) %>%
  print()
```

```{r plot-top50}
top50 <- sub_routes %>%
  filter(!is.na(gravity)) %>%
  slice_max(gravity, n = 50)

top50_sf <- top50 %>%
  rowwise() %>%
  mutate(geometry = st_sfc(
    st_linestring(matrix(c(src_lon, src_lat, dst_lon, dst_lat), ncol = 2, byrow = TRUE)),
    crs = 4326
  )) %>%
  ungroup() %>%
  st_as_sf() %>%
  st_transform(3857)

ap50_ids <- union(top50$src_id, top50$dst_id)
ap50 <- airport_access %>%
  filter(id %in% ap50_ids) %>%
  st_transform(3857)

# Label midpoints for the top 10 routes
top10_routes <- top50_sf %>% slice_head(n = 10) %>%
  mutate(
    mid = st_line_sample(geometry, sample = 0.5),
    label = paste0(src_iata, "–", dst_iata)
  )

p_top50 <- ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
  geom_sf(data = top50_sf, aes(color = rail_time_h, linewidth = log10(gravity)),
          alpha = 0.75) +
  geom_sf(data = ap50, color = C_APHSR, size = 3, alpha = 0.8) +
  geom_sf_text(data = top10_routes, aes(label = label, geometry = mid),
               size = 2.2, fontface = "bold", color = "#222222") +
  scale_color_gradientn(
    colours = c("#1a9850", "#91cf60", "#fee08b", "#fc8d59", "#d73027"),
    name    = "Gravity score  (pax_origin × pax_dest)",
    guide   = guide_colorbar(title = "Rail travel time (hours)", title.position = "top", barwidth = 15)
  ) +
  scale_linewidth_continuous(range = c(0.4, 3), guide = "none") +
  labs(
    title    = "Top 50 Short-Haul Routes with Rail-Competitive Travel Times",
    subtitle = "Both airports ≤ 5 km from HSR  |  Flight ≤ 2.5 h  |  Ranked by passenger gravity",
    caption  = SOURCE
  ) +
  theme_map()

ggsave(file.path(root_dir, "slides", "07_top50_substitutable.png"),
       p_top50, width = 14, height = 8.75, dpi = 150, bg = "white")
p_top50
```

---

## 7. Carbon-Adjusted Market Access

Following Allen & Arkolakis (2014), each airport's market access is:

$$MA_i = \sum_{j \neq i} Y_j \cdot \tau_{ij}^{-\theta}$$

where $Y_j$ is annual passengers (Eurostat), $\tau_{ij}$ is the
carbon-adjusted bilateral cost, and $\theta$ is the travel elasticity. The
bilateral cost embeds both time and carbon:

$$\tau_{ij} = t_{ij} + \lambda \cdot \text{CO}_2^{ij}
  \quad \text{where} \quad
  \lambda = \frac{\text{carbon price}}{\text{value of time}}
           = \frac{\text{EUR}\,100/\text{t}}{\text{EUR}\,30/\text{hr}}$$

The counterfactual replaces aviation cost with HSR cost on substitutable
routes. $\Delta MA_i$ measures the resulting change in connectivity.

**Parameter choices:**

- Aviation CO₂: 255 g/pax-km (includes radiative forcing)
- HSR CO₂: 14 g/pax-km (European average)
- θ tested at 3, 5, 8 — direction is robust, magnitude is sensitive

### Limitations

- **θ calibration** — borrowed from the trade literature (4–8 range), not
  estimated from transport data. Results shift a lot with this choice.
- **λ parameters** — carbon price of EUR 100/t is above current EU ETS
  (~EUR 60–70/t). Value of time of EUR 30/hr is a population average that
  misses the business vs. leisure split.
- **CO₂ accounting** — radiative forcing multiplier (1.9–2.9×) may or may not
  be included. If not, aviation emissions are understated.
- **$Y_j$ as demand proxy** — passenger volumes don't track economic mass well.
  A small hub serving a high-income catchment is undercounted vs. a large
  leisure hub.
- **Full substitution assumed** — the counterfactual swaps aviation cost
  entirely for HSR. Real modal shift is partial and gradual.
- **Additive cost function** — assumes time and carbon trade off linearly at a
  fixed rate.
- **Static model** — no induced demand, no frequency response, no general
  equilibrium effects.

```{r market-access}
HSR_SPEED    <- RAIL_SPEED
CO2_FLIGHT   <- 255
CO2_RAIL     <- 14
CARBON_PRICE <- 0.10   # EUR/kg (= EUR 100/tonne)
VOT          <- 30     # EUR/hr
LAMBDA       <- CARBON_PRICE / VOT
THETAS       <- c(3, 5, 8)

ap_universe <- europe_civil_pax %>%
  filter(!is.na(passengers)) %>%
  distinct(id, .keep_all = TRUE)

ap_ids <- ap_universe$id
ap_pax <- setNames(ap_universe$passengers, ap_universe$id)

routes_ma <- eu_routes %>%
  filter(src_id %in% ap_ids, dst_id %in% ap_ids) %>%
  distinct(src_id, dst_id, .keep_all = TRUE) %>%
  left_join(
    airport_access_lookup %>%
      rename(src_id = id, src_access_time_h = access_time_h),
    by = "src_id"
  ) %>%
  left_join(
    airport_access_lookup %>%
      rename(dst_id = id, dst_access_time_h = access_time_h),
    by = "dst_id"
  ) %>%
  mutate(
    dist_km       = haversine_km(src_lat, src_lon, dst_lat, dst_lon),
    flight_time_h = dist_km / 800 + 0.75,
    rail_time_h   = dist_km / RAIL_SPEED + src_access_time_h + dst_access_time_h + STATION_PENALTY,
    cost_baseline = flight_time_h + CO2_FLIGHT * dist_km / 1000 * LAMBDA,
    cost_counter  = cost_baseline,
    route_key     = map2_chr(src_id, dst_id,
                             ~ paste(sort(c(.x, .y)), collapse = "_"))
  )

sub_keys <- sub_routes$route_key
routes_ma <- routes_ma %>%
  mutate(cost_counter = if_else(
    route_key %in% sub_keys,
    rail_time_h + CO2_RAIL * dist_km / 1000 * LAMBDA,
    cost_counter
  )) %>%
  mutate(Y_dst = ap_pax[as.character(dst_id)])

ap_df <- ap_universe

for (theta in THETAS) {
  cb_col <- paste0("cb_", theta)
  cc_col <- paste0("cc_", theta)

  routes_ma[[cb_col]] <- routes_ma$Y_dst * routes_ma$cost_baseline^(-theta)
  routes_ma[[cc_col]] <- routes_ma$Y_dst * routes_ma$cost_counter^(-theta)

  ma_base    <- routes_ma %>% group_by(src_id) %>%
    summarise(val = sum(.data[[cb_col]], na.rm = TRUE), .groups = "drop")
  ma_counter <- routes_ma %>% group_by(src_id) %>%
    summarise(val = sum(.data[[cc_col]], na.rm = TRUE), .groups = "drop")

  ap_df[[paste0("MA_base_",      theta)]] <- ma_base$val[   match(ap_df$id, ma_base$src_id)]
  ap_df[[paste0("MA_counter_",   theta)]] <- ma_counter$val[match(ap_df$id, ma_counter$src_id)]
  ap_df[[paste0("delta_MA_",     theta)]] <-
    (ap_df[[paste0("MA_counter_", theta)]] - ap_df[[paste0("MA_base_", theta)]]) /
     ap_df[[paste0("MA_base_", theta)]] * 100
  ap_df[[paste0("abs_delta_MA_", theta)]] <-
    ap_df[[paste0("MA_counter_", theta)]] - ap_df[[paste0("MA_base_", theta)]]
}

ap_gdf <- ap_df %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(3857)

cat("Top 10 by % ΔMA (θ=5):\n")
ap_df %>% slice_max(delta_MA_5, n = 10) %>%
  select(name, iata, delta_MA_3, delta_MA_5, delta_MA_8) %>%
  mutate(across(starts_with("delta"), ~ round(., 1))) %>%
  print()

cat("\nTop 10 by absolute ΔMA (θ=5):\n")
ap_df %>% slice_max(abs_delta_MA_5, n = 10) %>%
  select(name, iata, abs_delta_MA_3, abs_delta_MA_5, abs_delta_MA_8) %>%
  print()
```

```{r plot-delta-ma}
vmax5 <- quantile(abs(ap_gdf$delta_MA_5), 0.95, na.rm = TRUE)

p_dma <- ggplot() +
  annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
  geom_sf(data = ap_gdf, aes(color = delta_MA_5), size = 3, alpha = 0.85) +
  geom_text_repel(
    data  = ap_gdf %>% slice_max(delta_MA_5, n = 5),
    aes(label = iata, geometry = geometry),
    stat  = "sf_coordinates",
    size  = 2.5, fontface = "bold", color = "#1a6600",
    min.segment.length = 0
  ) +
  geom_text_repel(
    data  = ap_gdf %>% slice_min(delta_MA_5, n = 3),
    aes(label = iata, geometry = geometry),
    stat  = "sf_coordinates",
    size  = 2.5, fontface = "bold", color = "#8b0000",
    min.segment.length = 0
  ) +
  scale_color_gradient2(
    low      = "#d73027", mid = "#ffffbf", high = "#1a9850",
    midpoint = 0, limits = c(-vmax5, vmax5),
    name     = "ΔMA (%)  —  change in carbon-adjusted market access",
    guide    = guide_colorbar(title.position = "top", barwidth = 15)
  ) +
  labs(
    title    = "Change in Carbon-Adjusted Market Access  (θ = 5)",
    subtitle = "Allen & Arkolakis (2014)  |  Counterfactual: HSR substitution on eligible routes",
    caption  = SOURCE
  ) +
  theme_map()

ggsave(file.path(root_dir, "slides", "08_delta_MA_theta5.png"),
       p_dma, width = 14, height = 8.75, dpi = 150, bg = "white")
p_dma
```

```{r plot-theta-sensitivity, fig.width=21, fig.height=8}
make_dma_panel <- function(theta) {
  col  <- paste0("delta_MA_", theta)
  vmax <- quantile(abs(ap_gdf[[col]]), 0.95, na.rm = TRUE)
  ggplot() +
    annotation_map_tile(type = "cartolight", zoom = 4, quiet = TRUE) +
    geom_sf(data = ap_gdf, aes(color = .data[[col]]),
            size = 2.5, alpha = 0.85) +
    scale_color_gradient2(
      low = "#d73027", mid = "#ffffbf", high = "#1a9850",
      midpoint = 0, limits = c(-vmax, vmax),
      name  = "ΔMA (%)",
      guide = guide_colorbar(title.position = "top", barwidth = 8)
    ) +
    labs(title = paste("θ =", theta), caption = SOURCE) +
    theme_map()
}

p_sens <- map(THETAS, make_dma_panel) %>%
  wrap_plots(nrow = 1) +
  plot_annotation(
    title = "Sensitivity to Trade Elasticity θ  —  Change in Carbon-Adjusted Market Access",
    theme = theme(plot.title = element_text(face = "bold", size = 14))
  )

ggsave(file.path(root_dir, "slides", "09_theta_sensitivity.png"),
       p_sens, width = 21, height = 8, dpi = 150, bg = "white")
p_sens
```

---

## 8. Sensitivity Analysis and Policy Application

The two main parameters in the model — carbon price and airport overhead — map
directly to policy levers. I use Barcelona–Madrid as the case study since it's
the best-documented air-rail competition example in Europe.

**Key result:** At baseline (0.75 h overhead), BCN–MAD needs a carbon price of
around EUR 150–240/t to flip to HSR. At a more realistic overhead of 1 h, it
flips at anything above EUR 100/t.

**Notes:**

- BCN–MAD saw roughly 60% air passenger reduction after AVE opened in 2008
- Reducing airport overhead is probably the more tractable lever — it's an
  infrastructure problem, not a tax policy problem

```{r breakeven-curve, fig.width=12, fig.height=6.5}
median_access_pair_h <- median(sub_routes$src_access_time_h + sub_routes$dst_access_time_h, na.rm = TRUE)
distances    <- seq(50, 1400, length.out = 500)
costs_flight <- distances / 800 + 0.75 + CO2_FLIGHT * distances / 1000 * LAMBDA
costs_hsr    <- distances / RAIL_SPEED + median_access_pair_h + STATION_PENALTY +
                CO2_RAIL * distances / 1000 * LAMBDA
breakeven    <- median_access_pair_h / ((CO2_FLIGHT - CO2_RAIL) / 1000 * LAMBDA)

cat(sprintf("Break-even distance (baseline params): %.0f km\n", breakeven))
cat(sprintf("Routes below break-even: %d\n", sum(sub_routes$dist_km < breakeven,  na.rm = TRUE)))
cat(sprintf("Routes above break-even: %d\n", sum(sub_routes$dist_km >= breakeven, na.rm = TRUE)))

cost_df <- data.frame(
  dist       = rep(distances, 2),
  cost       = c(costs_flight, costs_hsr),
  mode       = rep(c("Short-haul flight  (time + carbon cost)",
                     "High-speed rail  (time + carbon cost)"), each = 500)
)

fill_df <- data.frame(
  dist       = distances,
  cost_f     = costs_flight,
  cost_r     = costs_hsr,
  hsr_better = costs_hsr < costs_flight
)

p_breakeven <- ggplot(cost_df, aes(dist, cost, color = mode)) +
  geom_ribbon(
    data = fill_df %>% filter( hsr_better),
    aes(x = dist, ymin = cost_r, ymax = cost_f),
    inherit.aes = FALSE, fill = C_APHSR, alpha = 0.12
  ) +
  geom_ribbon(
    data = fill_df %>% filter(!hsr_better),
    aes(x = dist, ymin = cost_f, ymax = cost_r),
    inherit.aes = FALSE, fill = C_HSR, alpha = 0.12
  ) +
  geom_line(linewidth = 2) +
  scale_color_manual(values = c(C_HSR, C_APHSR), name = NULL) +
  geom_vline(xintercept = breakeven, color = "#333333", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = 482.93,   color = "#f4a261", linewidth = 1.2, linetype = "dotted") +
  annotate("text", x = breakeven + 20, y = min(costs_hsr) + 0.05,
           label = sprintf("Break-even\n~%.0f km", breakeven),
           size = 3, color = "#333333") +
  annotate("text", x = 482.93 + 20, y = min(costs_hsr) + 0.05,
           label = "BCN–MAD\n483 km",
           size = 3, color = "#f4a261", fontface = "bold") +
  labs(
    x       = "Route distance (km)",
    y       = "Carbon-adjusted travel cost (hours)",
    title   = "Flight vs HSR: Carbon-Adjusted Cost by Distance",
    subtitle = "Carbon = €100/t  |  Overhead = 0.75 h  |  HSR = 250 km/h avg",
    caption = SOURCE
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title      = element_text(face = "bold")
  )

ggsave(file.path(root_dir, "slides", "10_breakeven_curve.png"),
       p_breakeven, width = 12, height = 6.5, dpi = 150, bg = "white")
p_breakeven
```

```{r bcn-mad-heatmap, fig.width=12, fig.height=7}
carbon_prices <- c(50, 75, 100, 130, 150, 200, 300)
overheads     <- c(0.75, 1.0, 1.25, 1.5, 1.75, 2.0)
dist_bm       <- 482.93

heatmap_df <- expand.grid(overhead = overheads, carbon_price = carbon_prices) %>%
  mutate(
    lam     = carbon_price / (1000 * VOT),
    cf      = dist_bm / 800 + overhead + CO2_FLIGHT * dist_bm / 1000 * lam,
    cr      = dist_bm / RAIL_SPEED + median_access_pair_h + STATION_PENALTY +
              CO2_RAIL * dist_bm / 1000 * lam,
    hsr_win = cr < cf,
    margin  = (cf - cr) * 60   # minutes
  )

p_heatmap <- ggplot(heatmap_df,
                    aes(x = factor(carbon_price), y = factor(overhead),
                        fill = hsr_win)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%s%.0f min",
                                ifelse(margin > 0, "+ ", ""), margin)),
            size = 3.5, fontface = "bold",
            color = ifelse(abs(heatmap_df$margin) > 9, "white", "#333333")) +
  scale_fill_manual(
    values = c("FALSE" = "#d73027", "TRUE" = "#1a9850"),
    labels = c("FALSE" = "Flight preferred", "TRUE" = "HSR preferred"),
    name   = NULL
  ) +
  geom_vline(xintercept = c(2.5, 3.5), color = "#003399",
             linewidth = c(1.5, 1), linetype = c("dashed", "dotted")) +
  annotate("text", x = 2.6, y = 0.4, label = "Current ETS\n~€130",
           color = "#003399", size = 2.8, fontface = "bold", hjust = 0) +
  annotate("text", x = 3.6, y = 0.4, label = "2030 target",
           color = "#003399", size = 2.8, hjust = 0) +
  geom_hline(yintercept = 2.5, color = "#555555", linewidth = 1,
             linetype = "dashed") +
  annotate("text", x = 7.4, y = 2.4, label = "Realistic overhead",
           color = "#555555", size = 2.8, hjust = 1) +
  labs(
    x        = "Carbon price  (€/tonne CO₂)",
    y        = "Airport boarding overhead\n(security + boarding + taxi)",
    title    = "Barcelona–Madrid (483 km): When Does HSR Dominate?",
    subtitle = "Cell values = minutes saved (+) or lost (−) by switching to rail",
    caption  = SOURCE
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold"),
    legend.position = "top"
  )

ggsave(file.path(root_dir, "slides", "11_bcn_mad_heatmap.png"),
       p_heatmap, width = 12, height = 7, dpi = 150, bg = "white")
p_heatmap
```

---

## 9. Summary

**Main findings:**

- HSR is cost-competitive up to ~385 km at EUR 100/t carbon price and 0.75 h overhead
- 34 of 187 substitutable routes fall below that threshold
- BCN–MAD needs either ~EUR 150/t carbon price or overhead cut to ~1 h to flip
  — both plausible under current policy
- LHR: 0 substitutable routes. CDG: 21. Airport–rail integration makes a big
  difference.

**Policy takeaway:**

- Airport–rail integration is the more actionable near-term lever
- Blanket short-haul bans don't work — the spatial heterogeneity is too large,
  route-level policy is needed

**Limitations:**

- 5 km buffer overstates HSR access for airports without direct terminal
  connections
- Gravity score is a rough demand proxy — actual O-D data would improve route
  ranking
- No demand response or induced traffic effects modelled

**References:**

- Allen & Arkolakis (2014). Trade and the Topography of the Spatial Economy.
  *QJE* 129(3).
- Givoni & Dobruszkes (2013). A Review of Ex-Post Evidence for Mode
  Substitution. *Transport Reviews* 33(6).
- European Environment Agency (2023). *Transport and Environment Report*.
- Eurostat (2025). Air Transport Statistics: `avia_paoa`.

---

## Appendix: 4-Hour Reachability Isochrones

```{r isochrone-function}
plot_isochrone <- function(hub_iata, time_budget = 4.0) {

  hub_row  <- europe_civil %>% filter(iata == hub_iata) %>% slice(1)
  hub_lat  <- hub_row$lat
  hub_lon  <- hub_row$lon

  ap <- europe_civil %>%
    mutate(
      dist_km       = haversine_km(hub_lat, hub_lon, lat, lon),
      flight_time_h = dist_km / 800 + 0.75,
      hsr_time_h    = dist_km / RAIL_SPEED +
                      airport_access_lookup$access_time_h[match(hub_row$id, airport_access_lookup$id)] +
                      airport_access_lookup$access_time_h[match(id, airport_access_lookup$id)] +
                      STATION_PENALTY
    )

  by_flight <- ap %>% filter(flight_time_h <= time_budget)
  by_hsr    <- ap %>% filter(hsr_time_h <= time_budget)

  ap <- ap %>%
    mutate(reach = case_when(
      id %in% by_flight$id & id %in% by_hsr$id  ~ "Both modes",
      id %in% by_flight$id                       ~ "Flight only",
      id %in% by_hsr$id                          ~ "HSR only",
      TRUE                                        ~ "Beyond budget"
    ))

  pal   <- c("Flight only"    = C_HSR,
             "Both modes"     = "#9b72cf",
             "HSR only"       = C_APHSR,
             "Beyond budget"  = C_APNONE)
  sizes <- c("Flight only" = 2.5, "Both modes" = 2.5,
             "HSR only"    = 2.5, "Beyond budget" = 0.8)
  alpha <- c("Flight only" = 0.85, "Both modes" = 0.85,
             "HSR only"    = 0.85, "Beyond budget" = 0.2)

  ap_sf <- ap %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
    st_transform(3857)

  hub_sf <- hub_row %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
    st_transform(3857)

  # Convex hulls for flight and HSR reach
  hull_flight <- by_flight %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
    st_union() %>% st_convex_hull() %>% st_transform(3857)

  hull_hsr <- by_hsr %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
    st_union() %>% st_convex_hull() %>% st_transform(3857)

  counts <- ap %>% filter(reach != "Beyond budget") %>% count(reach)

  ggplot() +
    annotation_map_tile(type = "cartolight", zoom = 5, quiet = TRUE) +
    geom_sf(data = hull_flight, fill = C_HSR,   color = C_HSR,
            alpha = 0.07, linewidth = 1, linetype = "dashed") +
    geom_sf(data = hull_hsr,   fill = C_APHSR, color = C_APHSR,
            alpha = 0.07, linewidth = 1, linetype = "dashed") +
    geom_sf(data = ap_sf,
            aes(color = reach, size = reach, alpha = reach)) +
    geom_sf(data = hub_sf, shape = 9, size = 6, color = "#111111") +
    geom_sf_label(data = hub_sf, aes(label = iata),
                  size = 4, fontface = "bold", nudge_x = 50000) +
    scale_color_manual(values = pal,  name = NULL) +
    scale_size_manual( values = sizes, name = NULL) +
    scale_alpha_manual(values = alpha, name = NULL) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    labs(caption = SOURCE) +
    theme_map()
}
```

```{r isochrone-bcn}
p_bcn <- plot_isochrone("BCN", time_budget = 4.0) +
  labs(
    title    = "4h Reachability from Barcelona BCN  —  Flight vs HSR",
    subtitle = "Pyrenees constrain HSR reach northward; most central European cities remain flight-only"
  )

ggsave(file.path(root_dir, "slides", "17_isochrone_bcn.png"),
       p_bcn, width = 14, height = 8.75, dpi = 150, bg = "white")
p_bcn
```

```{r isochrone-fra}
p_fra <- plot_isochrone("FRA", time_budget = 4.0) +
  labs(
    title    = "4h Reachability from Frankfurt FRA  —  Flight vs HSR",
    subtitle = "Direct ICE airport station connects FRA to Paris, Brussels, Amsterdam, Basel and beyond"
  )

ggsave(file.path(root_dir, "slides", "18_isochrone_fra.png"),
       p_fra, width = 14, height = 8.75, dpi = 150, bg = "white")
p_fra
```

```{r isochrone-comparison, fig.width=22, fig.height=8}
hubs <- list(
  list(iata = "BCN", note = "Iberian Peninsula  —  Pyrenees barrier limits rail reach north"),
  list(iata = "CDG", note = "Central hub  —  TGV terminal integrated, highest ΔMA in model"),
  list(iata = "FRA", note = "Germany  —  Direct ICE airport station, densest European HSR node")
)

p_comparison <- map(hubs, ~ {
  plot_isochrone(.x$iata, time_budget = 4.0) +
    labs(title = sprintf("%s  —  %s", .x$iata, .x$note))
}) %>%
  wrap_plots(nrow = 1) +
  plot_annotation(
    title = "4h Reachability by Mode  —  BCN vs CDG vs FRA",
    subtitle = "HSR reach approximated via great-circle distance, restricted to airports with HSR access",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 10)
    )
  )

ggsave(file.path(root_dir, "slides", "19_isochrone_comparison.png"),
       p_comparison, width = 22, height = 8, dpi = 150, bg = "white")
p_comparison
```
