---
title: "Western Europe Hub Airports: HSR Substitution Suitability"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

This notebook applies the Barcelona route-flagging approach to the current 15
largest western-European airports in the repo. For each outbound route in
`data/clean/master_routes_clean.csv`, it estimates a rail travel time using the
GL2017 passenger rail network, then flags routes as plausible HSR alternatives
when they are:

- connector-feasible
- rail-connected
- below a simple rail-time threshold

The airport-level "carbon suitability" score is defined here as the share of
observed outbound passengers on flagged routes. That is a simple, interpretable
proxy for where a shift from short flights to rail could affect the most people.

**Research question:** Among the 15 main western-European airports in this
project, which hubs appear most suitable for substituting outbound short-haul
flights with rail, once we use observed route counts and passengers rather than
the earlier gravity proxy?

**Approach:**

- Build a route-level rail-time model on the GL2017 passenger rail network
- Apply a simple HSR-alternative flag to observed outbound routes
- Summarise route, flight, and passenger substitution potential by airport
- Map airport-level substitution suitability across the 15 hubs

**Data:** GL2017 rail, cleaned airport layers, and Eurostat-derived route and
airport totals already built in this repo

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 13,
  fig.height = 8
)
```

## 1. Shared Setup

```{r packages}
suppressPackageStartupMessages({
  library(sf)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(purrr)
  library(stringr)
  library(ggplot2)
  library(scales)
  library(ggrepel)
  library(here)
  library(sfnetworks)
  library(tidygraph)
  library(units)
})

source(here::here("scripts", "paths.R"))
```

```{r constants}
ANALYSIS_CRS <- 3035
C_HSR <- "#e63946"
C_CONV <- "#cccccc"
C_APHSR <- "#2176ae"
C_APNONE <- "#aaaaaa"
SOURCE <- "Data: GL2017 rail, cleaned airport layers, Eurostat-derived route totals"
PASSENGER_RAIL_CLASSES <- c("Passenger", "Passenger and freight")
HSR_SPEED_KMH <- 250
CONVENTIONAL_SPEED_KMH <- 110
CONNECTOR_SPEED_KMH <- 45
MAX_CONNECTOR_KM <- 25
MIN_CONTIGUOUS_HSR_KM <- 50
TRANSFER_PENALTY_H <- 0.5
ALTERNATIVE_TIME_H <- 6

routes_path <- file.path(data_clean, "master_routes_clean.csv")
airports_path <- file.path(data_clean, "all_airports.gpkg")
main_airports_path <- file.path(data_clean, "main_airports.gpkg")
rail_path <- file.path(data_clean, "shapefiles", "railways_GL2017_EU.shp")
countries_path <- file.path(data_clean, "ne_10m_countries.gpkg")

required_paths <- c(routes_path, airports_path, main_airports_path, rail_path, countries_path)
stopifnot(all(file.exists(required_paths)))

theme_map <- function(...) {
  theme_void(...) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0),
      plot.subtitle = element_text(size = 10, hjust = 0, color = "#444444"),
      plot.caption = element_text(size = 7.5, color = "#999999", face = "italic", hjust = 1),
      legend.position = "bottom"
    )
}
```

## 2. Airport and Route Data

```{r read-inputs}
routes_raw <- read_csv(routes_path, show_col_types = FALSE)

airports_raw <- st_read(airports_path, quiet = TRUE) %>%
  st_make_valid() %>%
  st_transform(ANALYSIS_CRS)

main_airports_raw <- st_read(main_airports_path, quiet = TRUE) %>%
  st_make_valid() %>%
  st_transform(ANALYSIS_CRS)

rail_raw <- st_read(rail_path, quiet = TRUE) %>%
  st_make_valid() %>%
  st_transform(ANALYSIS_CRS)

countries <- st_read(countries_path, quiet = TRUE) %>%
  st_make_valid() %>%
  st_transform(ANALYSIS_CRS)
```

```{r audit}
cat("Routes in master_routes_clean.csv:", nrow(routes_raw), "\n")
cat("Distinct route origins:", n_distinct(routes_raw$start_icao), "\n")
cat("Airport points in all_airports.gpkg:", nrow(airports_raw), "\n")
cat("Main airports in main_airports.gpkg:", nrow(main_airports_raw), "\n")
cat("GL2017 passenger-capable segments before filtering:", nrow(rail_raw), "\n")
```

## 3. Observed Passenger Demand and Route Tables

```{r clean-airports-routes}
safe_max <- function(x) {
  if (all(is.na(x))) {
    return(NA_real_)
  }
  max(x, na.rm = TRUE)
}

airports_clean <- airports_raw %>%
  mutate(
    station_like = str_detect(
      str_to_upper(AirportName),
      "RAILWAY STATION|\\bHBF\\b|\\bSTATION\\b"
    ),
    name_length = nchar(AirportName)
  ) %>%
  arrange(ICAO, station_like, desc(name_length)) %>%
  group_by(ICAO) %>%
  slice(1) %>%
  ungroup() %>%
  select(-name_length)

main_airports <- main_airports_raw %>%
  mutate(
    station_like = str_detect(
      str_to_upper(AirportName),
      "RAILWAY STATION|\\bHBF\\b|\\bSTATION\\b"
    ),
    name_length = nchar(AirportName)
  ) %>%
  arrange(ICAO, station_like, desc(name_length)) %>%
  group_by(ICAO) %>%
  slice(1) %>%
  ungroup() %>%
  select(-name_length)

route_pairs <- routes_raw %>%
  group_by(start_icao, end_icao) %>%
  summarise(
    start_airport = first(start_airport),
    start_iata = first(start_iata),
    start_country_name = first(start_country_name),
    start_country_a2 = first(start_country_a2),
    end_airport = first(end_airport),
    end_iata = first(end_iata),
    end_country_name = first(end_country_name),
    end_country_a2 = first(end_country_a2),
    route_name = first(route_name),
    total_flights = safe_max(total_flights),
    route_passengers = safe_max(route_passengers),
    start_pax_total = safe_max(start_pax_total),
    end_pax_total = safe_max(end_pax_total),
    start_pax_pct = safe_max(start_pax_pct),
    reporting_country = first(reporting_country),
    year = first(year),
    destination_in_western_europe = first(destination_in_western_europe),
    .groups = "drop"
  )

route_airport_codes <- union(route_pairs$start_icao, route_pairs$end_icao)
route_airports <- airports_clean %>%
  filter(ICAO %in% route_airport_codes) %>%
  transmute(ICAO, AirportName)
```

## 4. European Rail Infrastructure and Access Model

```{r prepare-rail}
rail_passenger <- rail_raw %>%
  filter(RAILWAYS_A %in% PASSENGER_RAIL_CLASSES) %>%
  filter(!st_is_empty(geometry), !is.na(geometry)) %>%
  st_collection_extract("LINESTRING") %>%
  mutate(
    speed_kmh = if_else(TYPE == "High speed", HSR_SPEED_KMH, CONVENTIONAL_SPEED_KMH),
    rail_class = if_else(TYPE == "High speed", "High speed", "Conventional")
  )

cat("Passenger-capable GL2017 rail segments retained:", nrow(rail_passenger), "\n")
print(rail_passenger %>% st_drop_geometry() %>% count(TYPE, sort = TRUE))

rail_bbox <- st_bbox(rail_passenger)
rail_xlim <- c(rail_bbox$xmin, rail_bbox$xmax)
rail_ylim <- c(rail_bbox$ymin, rail_bbox$ymax)
```

### Rail Network and Main Airports

```{r rail-network-map, fig.cap="GL2017 passenger rail network and the 15 main airports"}
rail_network_map <- ggplot() +
  geom_sf(data = countries, fill = "#F8F8F6", color = "#D0D5D9", linewidth = 0.16) +
  geom_sf(
    data = rail_passenger %>% filter(TYPE != "High speed"),
    color = "#9FB88D",
    linewidth = 0.22,
    alpha = 0.45
  ) +
  geom_sf(
    data = rail_passenger %>% filter(TYPE == "High speed"),
    color = C_HSR,
    linewidth = 0.45,
    alpha = 0.85
  ) +
  geom_sf(
    data = main_airports,
    shape = 21,
    size = 3.6,
    fill = "#7DE3EA",
    color = "#143C54",
    stroke = 0.35
  ) +
  geom_text_repel(
    data = main_airports %>% cbind(st_coordinates(.)),
    aes(X, Y, label = IATA),
    size = 3.2,
    box.padding = 0.35,
    point.padding = 0.2,
    min.segment.length = 0
  ) +
  coord_sf(xlim = rail_xlim, ylim = rail_ylim, expand = FALSE) +
  labs(
    title = "GL2017 Passenger Rail Network and the 15 Main Airports",
    subtitle = "High-speed segments are highlighted in red; conventional passenger rail is shown in green",
    caption = SOURCE
  ) +
  theme_map()

rail_network_map
```

```{r snap-airports}
nearest_edge_idx <- st_nearest_feature(route_airports, rail_passenger)

snapped_geoms <- lapply(seq_len(nrow(route_airports)), function(i) {
  pair <- st_nearest_points(route_airports[i, ], rail_passenger[nearest_edge_idx[i], ])
  snapped_pair_points <- st_cast(pair, "POINT")
  snapped_pair_points[2]
})

snapped_sfc <- do.call(c, snapped_geoms)
st_crs(snapped_sfc) <- st_crs(ANALYSIS_CRS)

snapped_airports <- st_as_sf(
  route_airports %>% st_drop_geometry(),
  geometry = snapped_sfc
) %>%
  mutate(
    nearest_edge_idx = nearest_edge_idx,
    access_km = as.numeric(
      set_units(st_distance(route_airports, geometry, by_element = TRUE), "km")
    ),
    access_time_h = access_km / CONNECTOR_SPEED_KMH,
    within_connector_limit = access_km <= MAX_CONNECTOR_KM
  )

snapped_airports %>%
  st_drop_geometry() %>%
  slice_max(order_by = access_km, n = 20)
```

### Airport-to-Rail Accessibility

```{r airport-access-map-data}
main_airport_access <- main_airports %>%
  rename(ICAO = ICAO, IATA = IATA) %>%
  left_join(
    snapped_airports %>%
      st_drop_geometry() %>%
      select(ICAO, access_km, access_time_h, within_connector_limit),
    by = "ICAO"
  ) %>%
  mutate(
    access_status = if_else(within_connector_limit, "Within modeled rail catchment", "No nearby rail in dataset")
  )
```

```{r airport-access-map, fig.cap="Distance from each main airport to the nearest passenger rail segment"}
airport_access_map <- ggplot() +
  geom_sf(data = countries, fill = "#F8F8F6", color = "#D0D5D9", linewidth = 0.16) +
  geom_sf(data = rail_passenger, color = "#AEB8C2", linewidth = 0.18, alpha = 0.35) +
  geom_sf(
    data = main_airport_access %>% filter(within_connector_limit),
    aes(color = access_km, size = access_km),
    alpha = 0.9
  ) +
  geom_sf(
    data = main_airport_access %>% filter(!within_connector_limit),
    color = "#8F8F8F",
    size = 3.8,
    alpha = 0.95
  ) +
  geom_text_repel(
    data = main_airport_access %>% cbind(st_coordinates(.)),
    aes(X, Y, label = IATA),
    size = 3.2,
    box.padding = 0.35,
    point.padding = 0.2,
    min.segment.length = 0
  ) +
  scale_color_gradientn(
    colours = c("#1a9850", "#91cf60", "#fee08b", "#fc8d59", "#d73027"),
    name = "Distance to rail (km)"
  ) +
  scale_size_continuous(range = c(2.5, 6), guide = "none") +
  coord_sf(xlim = rail_xlim, ylim = rail_ylim, expand = FALSE) +
  labs(
    title = "Airport-to-Rail Accessibility for the 15 Main Airports",
    subtitle = paste0("Airports within ", MAX_CONNECTOR_KM, " km are treated as connector-feasible; grey points have no nearby rail in this dataset"),
    caption = SOURCE
  ) +
  theme_map()

airport_access_map
```

## 5. Spatial Analysis — HSR Airport Access

The flood-fill diagnostic below mirrors the older draft notebook. I keep it as
an optional reachability visualization rather than the main travel-time model:
an airport that sits within the connector threshold of one HSR segment
effectively gains access to the whole geometrically connected corridor.

### Optional Flood-Fill Diagnostic

```{r flood-fill-data}
hs_only <- rail_passenger %>%
  filter(TYPE == "High speed")

airport_buffers <- main_airports %>%
  st_buffer(MAX_CONNECTOR_KM * 1000)

accessible <- as.logical(st_intersects(hs_only, st_union(airport_buffers), sparse = FALSE))

for (i in seq_len(100)) {
  prev <- sum(accessible)
  expanded <- hs_only$geometry[accessible] %>% st_union() %>% st_buffer(200)
  accessible <- as.logical(st_intersects(hs_only, expanded, sparse = FALSE))
  if (sum(accessible) == prev) {
    break
  }
}

hs_accessible <- hs_only[accessible, ]
hs_not_accessible <- hs_only[!accessible, ]
main_airports_hsr_access <- main_airport_access %>%
  mutate(hsr_access = within_connector_limit)
```

```{r flood-fill-map, fig.cap="Optional diagnostic: connected GL2017 high-speed subnetwork reachable from the 15 airports"}
flood_fill_map <- ggplot() +
  geom_sf(data = countries, fill = "#F8F8F6", color = "#D0D5D9", linewidth = 0.16) +
  geom_sf(data = hs_not_accessible, color = "#F4A261", linewidth = 0.35, alpha = 0.55) +
  geom_sf(data = hs_accessible, color = "#D1495B", linewidth = 0.55, alpha = 0.9) +
  geom_sf(
    data = main_airports_hsr_access,
    aes(color = hsr_access),
    size = 2.6,
    alpha = 0.9
  ) +
  scale_color_manual(
    values = c("TRUE" = "#1B9E77", "FALSE" = "#7F8C8D"),
    labels = c("FALSE" = "Outside snap cap", "TRUE" = "Within snap cap"),
    name = NULL
  ) +
  coord_sf(xlim = rail_xlim, ylim = rail_ylim, expand = FALSE) +
  labs(
    title = "Reachable High-Speed Corridor System from the 15 Airports",
    subtitle = "Airports trigger access to connected GL2017 high-speed corridors within the snap threshold",
    caption = "This is a corridor-reachability visualization, not the route-specific travel-time model"
  ) +
  theme_map()

flood_fill_map
```

```{r build-network}
rail_net <- rail_passenger %>%
  as_sfnetwork(directed = FALSE) %>%
  convert(to_spatial_subdivision)

rail_net <- st_network_blend(rail_net, snapped_airports)

rail_net <- rail_net %>%
  activate("edges") %>%
  mutate(
    edge_length_km = as.numeric(set_units(st_length(geometry), "km"))
  )

network_nodes <- st_as_sf(rail_net, "nodes")
network_edges <- st_as_sf(rail_net, "edges")

snapped_airports <- snapped_airports %>%
  mutate(node_id = st_nearest_feature(geometry, network_nodes))
```

## 6. Substitutable Routes with Observed Counts

```{r route-function}
build_route_result <- function(origin_code, dest_code) {
  origin_info <- snapped_airports %>% filter(ICAO == origin_code)
  dest_info <- snapped_airports %>% filter(ICAO == dest_code)

  if (!nrow(origin_info) || !nrow(dest_info)) {
    return(
      tibble(
        start_icao = origin_code,
        end_icao = dest_code,
        connector_feasible = FALSE,
        rail_connected = FALSE,
        rail_km = NA_real_,
        raw_high_speed_km = NA_real_,
        high_speed_km = NA_real_,
        conventional_km = NA_real_,
        rail_in_vehicle_h = NA_real_,
        origin_access_km = NA_real_,
        destination_access_km = NA_real_,
        origin_access_h = NA_real_,
        destination_access_h = NA_real_,
        transfer_penalty_h = TRANSFER_PENALTY_H,
        rail_total_time_h = NA_real_
      )
    )
  }

  if (!isTRUE(origin_info$within_connector_limit) || !isTRUE(dest_info$within_connector_limit)) {
    return(
      tibble(
        start_icao = origin_code,
        end_icao = dest_code,
        connector_feasible = FALSE,
        rail_connected = FALSE,
        rail_km = NA_real_,
        raw_high_speed_km = NA_real_,
        high_speed_km = NA_real_,
        conventional_km = NA_real_,
        rail_in_vehicle_h = NA_real_,
        origin_access_km = origin_info$access_km,
        destination_access_km = dest_info$access_km,
        origin_access_h = origin_info$access_time_h,
        destination_access_h = dest_info$access_time_h,
        transfer_penalty_h = TRANSFER_PENALTY_H,
        rail_total_time_h = NA_real_
      )
    )
  }

  edge_weights <- network_edges$edge_length_km /
    ifelse(network_edges$TYPE == "High speed", HSR_SPEED_KMH, CONVENTIONAL_SPEED_KMH)

  path <- suppressWarnings(
    st_network_paths(
      rail_net,
      from = origin_info$node_id,
      to = dest_info$node_id,
      weights = edge_weights
    )
  )

  edge_ids <- unlist(path$edge_paths)

  if (!length(edge_ids)) {
    return(
      tibble(
        start_icao = origin_code,
        end_icao = dest_code,
        connector_feasible = TRUE,
        rail_connected = FALSE,
        rail_km = NA_real_,
        raw_high_speed_km = NA_real_,
        high_speed_km = NA_real_,
        conventional_km = NA_real_,
        rail_in_vehicle_h = NA_real_,
        origin_access_km = origin_info$access_km,
        destination_access_km = dest_info$access_km,
        origin_access_h = origin_info$access_time_h,
        destination_access_h = dest_info$access_time_h,
        transfer_penalty_h = TRANSFER_PENALTY_H,
        rail_total_time_h = NA_real_
      )
    )
  }

  path_edge_df <- network_edges[edge_ids, ] %>%
    st_drop_geometry() %>%
    mutate(
      is_hsr = TYPE == "High speed",
      run_id = cumsum(coalesce(is_hsr != lag(is_hsr), TRUE))
    )

  hsr_runs <- path_edge_df %>%
    group_by(run_id, is_hsr) %>%
    summarise(run_km = sum(edge_length_km), .groups = "drop")

  effective_high_speed_km <- hsr_runs %>%
    filter(is_hsr, run_km >= MIN_CONTIGUOUS_HSR_KM) %>%
    summarise(value = sum(run_km), .groups = "drop") %>%
    pull(value)

  if (!length(effective_high_speed_km) || is.na(effective_high_speed_km)) {
    effective_high_speed_km <- 0
  }

  raw_high_speed_km <- path_edge_df %>%
    filter(is_hsr) %>%
    summarise(value = sum(edge_length_km), .groups = "drop") %>%
    pull(value)

  if (!length(raw_high_speed_km) || is.na(raw_high_speed_km)) {
    raw_high_speed_km <- 0
  }

  rail_km <- sum(path_edge_df$edge_length_km)
  conventional_km <- rail_km - effective_high_speed_km

  rail_in_vehicle_h <- (effective_high_speed_km / HSR_SPEED_KMH) +
    (conventional_km / CONVENTIONAL_SPEED_KMH)

  tibble(
    start_icao = origin_code,
    end_icao = dest_code,
    connector_feasible = TRUE,
    rail_connected = TRUE,
    rail_km = rail_km,
    raw_high_speed_km = raw_high_speed_km,
    high_speed_km = effective_high_speed_km,
    conventional_km = conventional_km,
    rail_in_vehicle_h = rail_in_vehicle_h,
    origin_access_km = origin_info$access_km,
    destination_access_km = dest_info$access_km,
    origin_access_h = origin_info$access_time_h,
    destination_access_h = dest_info$access_time_h,
    transfer_penalty_h = TRANSFER_PENALTY_H,
    rail_total_time_h = rail_in_vehicle_h + origin_info$access_time_h + dest_info$access_time_h + TRANSFER_PENALTY_H
  )
}
```

```{r build-all-routes}
route_results <- map2_dfr(route_pairs$start_icao, route_pairs$end_icao, build_route_result)

all_routes <- route_pairs %>%
  left_join(route_results, by = c("start_icao", "end_icao")) %>%
  mutate(
    hsr_alternative = rail_connected & rail_total_time_h < ALTERNATIVE_TIME_H,
    counted_flights = coalesce(total_flights, 0),
    counted_passengers = coalesce(route_passengers, 0)
  )

cat("Total route pairs evaluated:", nrow(all_routes), "\n")
cat("Connector-feasible route pairs:", sum(all_routes$connector_feasible, na.rm = TRUE), "\n")
cat("Rail-connected route pairs:", sum(all_routes$rail_connected, na.rm = TRUE), "\n")
cat("Flagged HSR alternatives:", sum(all_routes$hsr_alternative, na.rm = TRUE), "\n")
```

### Route-Level Summary

```{r route-level-summary}
route_flag_summary <- all_routes %>%
  mutate(
    route_group = case_when(
      hsr_alternative ~ "Flagged as HSR alternative",
      rail_connected ~ "Rail-connected but slower (6 h+)",
      TRUE ~ "Not rail-feasible / disconnected"
    )
  ) %>%
  group_by(route_group) %>%
  summarise(
    routes = n(),
    flights = sum(counted_flights, na.rm = TRUE),
    passengers = sum(counted_passengers, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    route_pct = routes / sum(routes),
    flight_pct = flights / sum(flights),
    passenger_pct = passengers / sum(passengers)
  )

route_flag_summary
```

### Top Flagged Routes by Observed Passengers

```{r top-routes-map-data}
top_flagged_routes <- all_routes %>%
  filter(hsr_alternative, !is.na(route_passengers)) %>%
  arrange(desc(route_passengers)) %>%
  slice_head(n = 50)

top_route_airports <- airports_clean %>%
  filter(ICAO %in% union(top_flagged_routes$start_icao, top_flagged_routes$end_icao)) %>%
  transmute(
    ICAO,
    lon = GeoPointLong,
    lat = GeoPointLat,
    geometry = st_geometry(.)
  )

top_route_lines <- top_flagged_routes %>%
  left_join(top_route_airports %>% st_drop_geometry() %>% rename(start_icao = ICAO, start_lon = lon, start_lat = lat), by = "start_icao") %>%
  left_join(top_route_airports %>% st_drop_geometry() %>% rename(end_icao = ICAO, end_lon = lon, end_lat = lat), by = "end_icao") %>%
  rowwise() %>%
  mutate(
    geometry = st_sfc(
      st_linestring(
        matrix(c(start_lon, start_lat, end_lon, end_lat), ncol = 2, byrow = TRUE)
      ),
      crs = 4326
    )
  ) %>%
  ungroup() %>%
  st_as_sf() %>%
  st_transform(ANALYSIS_CRS)

top_route_airports_sf <- top_route_airports %>%
  filter(ICAO %in% union(top_flagged_routes$start_icao, top_flagged_routes$end_icao))
```

```{r top-routes-map, fig.cap="Top 50 flagged HSR-alternative routes ranked by observed passengers"}
top_routes_map <- ggplot() +
  geom_sf(data = countries, fill = "#F8F8F6", color = "#D0D5D9", linewidth = 0.16) +
  geom_sf(
    data = top_route_lines,
    aes(color = rail_total_time_h, linewidth = log10(route_passengers)),
    alpha = 0.72,
    lineend = "round"
  ) +
  geom_sf(
    data = top_route_airports_sf,
    color = "#1E2D3B",
    size = 2.1,
    alpha = 0.85
  ) +
  scale_color_gradientn(
    colours = c("#1a9850", "#91cf60", "#fee08b", "#fc8d59", "#d73027"),
    name = "Rail time (hours)"
  ) +
  scale_linewidth_continuous(range = c(0.35, 2.2), guide = "none") +
  coord_sf(expand = FALSE) +
  labs(
    title = "Top 50 Flagged HSR-Alternative Routes",
    subtitle = "Routes are ranked by observed passengers rather than a gravity proxy",
    caption = SOURCE
  ) +
  theme_map()

top_routes_map
```

## 7. Airport-Level Suitability Summary

```{r airport-summary}
airport_summary <- all_routes %>%
  group_by(start_airport, start_iata, start_icao, start_country_name, start_country_a2) %>%
  summarise(
    start_pax_total = safe_max(start_pax_total),
    outbound_routes = n(),
    outbound_flights = sum(counted_flights, na.rm = TRUE),
    outbound_passengers = sum(counted_passengers, na.rm = TRUE),
    alternative_routes = sum(hsr_alternative, na.rm = TRUE),
    alternative_flights = sum(counted_flights[hsr_alternative], na.rm = TRUE),
    alternative_passengers = sum(counted_passengers[hsr_alternative], na.rm = TRUE),
    route_share = alternative_routes / outbound_routes,
    flight_share = if_else(outbound_flights > 0, alternative_flights / outbound_flights, NA_real_),
    passenger_share = if_else(outbound_passengers > 0, alternative_passengers / outbound_passengers, NA_real_),
    .groups = "drop"
  ) %>%
  mutate(
    carbon_suitability_score = passenger_share
  ) %>%
  left_join(
    main_airport_access %>%
      st_drop_geometry() %>%
      transmute(start_iata = IATA, airport_access_km = access_km, within_connector_limit),
    by = "start_iata"
  ) %>%
  arrange(desc(carbon_suitability_score))

airport_summary
```

### Airport Suitability Map

```{r airport-map-data}
airport_map_data <- main_airports %>%
  rename(
    start_airport = AirportName,
    start_iata = IATA,
    start_icao = ICAO
  ) %>%
  left_join(airport_summary, by = c("start_airport", "start_iata", "start_icao"))

airport_map_data <- airport_map_data %>%
  mutate(
    score_status = if_else(within_connector_limit, "Scored from modeled rail access", "No nearby rail in dataset")
  )
```

```{r airport-suitability-map, fig.cap="Carbon-suitability proxy for the 15 main western-European airports"}
airport_suitability_map <- ggplot() +
  geom_sf(data = countries, fill = "#F8F8F6", color = "#D0D5D9", linewidth = 0.16) +
  geom_sf(
    data = airport_map_data %>% filter(within_connector_limit),
    aes(size = outbound_passengers, fill = carbon_suitability_score),
    shape = 21,
    color = "#1E2D3B",
    stroke = 0.35,
    alpha = 0.95
  ) +
  geom_sf(
    data = airport_map_data %>% filter(!within_connector_limit),
    shape = 21,
    size = 5.5,
    fill = "#BDBDBD",
    color = "#5F6368",
    stroke = 0.35,
    alpha = 0.98
  ) +
  geom_text_repel(
    data = airport_map_data %>%
      cbind(st_coordinates(.)),
    aes(X, Y, label = start_iata),
    size = 3.4,
    box.padding = 0.35,
    point.padding = 0.2,
    min.segment.length = 0
  ) +
  scale_fill_viridis_c(
    option = "cividis",
    labels = percent_format(accuracy = 1),
    name = "Passenger share\nflagged as HSR\nalternative"
  ) +
  scale_size_continuous(
    range = c(3, 11),
    labels = label_number(big.mark = ","),
    name = "Outbound passengers"
  ) +
  coord_sf(xlim = rail_xlim, ylim = rail_ylim, expand = FALSE) +
  labs(
    title = "Carbon Suitability Proxy Across the 15 Main Airports",
    subtitle = paste0(
      "Score = share of observed outbound passengers on routes flagged as HSR alternatives (rail-connected and < ",
      ALTERNATIVE_TIME_H,
      " h); grey airports sit outside the modeled rail catchment in this dataset"
    ),
    caption = SOURCE
  ) +
  theme_map()

airport_suitability_map
```

### Ranked Airport Scores

```{r ranked-airport-scores, fig.cap="Airport ranking by passenger-share substitution suitability"}
ranked_airport_scores <- airport_summary %>%
  mutate(start_iata = factor(start_iata, levels = rev(start_iata))) %>%
  ggplot(aes(x = carbon_suitability_score, y = start_iata, color = carbon_suitability_score)) +
  geom_segment(
    aes(x = 0, xend = carbon_suitability_score, y = start_iata, yend = start_iata),
    linewidth = 0.8,
    alpha = 0.7
  ) +
  geom_point(aes(size = outbound_passengers), alpha = 0.95) +
  scale_color_viridis_c(option = "cividis", labels = percent_format(accuracy = 1), guide = "none") +
  scale_size_continuous(range = c(2.5, 8), labels = label_number(big.mark = ","), name = "Outbound passengers") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Airport Ranking by HSR Substitution Suitability",
    subtitle = "Higher values mean a larger share of outbound passengers are on flagged HSR-alternative routes",
    x = "Passenger share flagged as HSR alternative",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

ranked_airport_scores
```

### Airport-Level Route / Flight / Passenger Shares

```{r airport-share-comparison, fig.cap="Comparison of route, flight, and passenger shares flagged as HSR alternatives"}
airport_share_long <- airport_summary %>%
  select(start_iata, route_share, flight_share, passenger_share) %>%
  pivot_longer(
    cols = c(route_share, flight_share, passenger_share),
    names_to = "metric",
    values_to = "share"
  ) %>%
  mutate(
    metric = recode(
      metric,
      route_share = "Route share",
      flight_share = "Flight share",
      passenger_share = "Passenger share"
    ),
    start_iata = factor(start_iata, levels = airport_summary$start_iata[order(airport_summary$carbon_suitability_score)])
  )

airport_share_comparison <- ggplot(airport_share_long, aes(x = start_iata, y = share, fill = metric)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.72) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(
    values = c(
      "Route share" = "#7B9ACC",
      "Flight share" = "#F2B134",
      "Passenger share" = "#2C7C5A"
    ),
    name = NULL
  ) +
  labs(
    title = "How the HSR Alternative Flag Scales by Airport",
    subtitle = "Comparing the share of routes, flights, and passengers flagged as substitutable by rail",
    x = NULL,
    y = "Share flagged as HSR alternative"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

airport_share_comparison
```

## 8. Summary and Export Tables

```{r export, eval=FALSE}
write_csv(all_routes, file.path(outputs, "all_15_airports_hsr_route_flags.csv"))
write_csv(airport_summary, file.path(outputs, "all_15_airports_hsr_airport_summary.csv"))
```

## 9. Summary

- The route flag is intentionally simple: rail-connected and below a 6-hour
  threshold after connector access and transfer penalty.
- The airport score is not a full welfare model. It is a descriptive proxy for
  where replacing outbound short flights with rail could affect the largest
  share of passengers.
- Because all routes share the same rail model assumptions, the scores are best
  interpreted comparatively across airports rather than as precise forecasts.
